---
title: "JAVELIN RCC Delayed Recurrence"
author: "Steven Monda"
date: "2026-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache=TRUE, message = FALSE, warning = FALSE,eval=TRUE)
```
## Load Necessary Packages
```{r packages, eval=T}
library(haven)
library(limma)
library(ggbeeswarm)
library(survival)
library(randomForest)
library(ggplot2)
library(matrixStats)
library(survminer)
library(readxl)
library(readr)
# library(umap)

library(reshape2)
library(dplyr)
library(gtsummary)
library(reshape2)
library(ggplot2)
library(patchwork)
library(patchwork)
library(cowplot)
library(scales)
library(rms)
library(gtsummary)
library(binom)
library(ComplexHeatmap)
library(circlize)
library(fgsea)
library(msigdbr)

```

## Data Harmonization
```{r import clinical data}
response<-read_sas("/Users/smonda/Library/CloudStorage/OneDrive-MichiganMedicine/Research/M vs P/RDS121_2024_12_12T10_04_unzip/adrs.sas7bdat")

medhx<-read_sas("/Users/smonda/Library/CloudStorage/OneDrive-MichiganMedicine/Research/M vs P/RDS121_2024_12_12T10_04_unzip/admh.sas7bdat")

clinbroad<-read_sas("/Users/smonda/Library/CloudStorage/OneDrive-MichiganMedicine/Research/M vs P/RDS121_2024_12_12T10_04_unzip/adsl.sas7bdat")

tte<-read_sas("/Users/smonda/Library/CloudStorage/OneDrive-MichiganMedicine/Research/M vs P/RDS121_2024_12_12T10_04_unzip/adtte.sas7bdat")

#clin broad then subsetted ot only patients that were randomiezed

clin<-subset(clinbroad,clinbroad$ARM %in% c("Avelumab 10 mg/kg IV Q2W in a 6-week cycle + axitinib 5 mg PO BID","Sunitinib 50 mg PO QD on Schedule 4/2"))
rm(clinbroad)

#again subsets medhx to only patients on trial and their cancer hx
cancerhx<-subset(medhx,
                 medhx$MHCAT=="PRIMARY DIAGNOSIS"&
                 medhx$USUBJID %in% clin$USUBJID)

cancerhx$age<-sub("-.*", "", tte$AGE[match(cancerhx$USUBJID,tte$USUBJID)])
cancerhx$pdl1<-tte$PDL1FL[match(cancerhx$USUBJID,tte$USUBJID)]

cancerhx$orr<-    subset(response$AVALCAT1,response$PARAMCD=="BOR_C")[match(cancerhx$USUBJID,subset(response$USUBJID,response$PARAMCD=="BOR_C"))]

#Imports processed log2(TPM+1) data from PMID: 39789182 https://pubmed.ncbi.nlm.nih.gov/39789182/
natjavrna <- read_excel("/Users/smonda/Library/CloudStorage/OneDrive-MichiganMedicine/Research/M vs P/RDS121_2024_12_12T10_04_unzip/43018_2024_896_MOESM2_ESM(4).xlsx", 
                        sheet = "Supp. Table 3")

natjavclin <-read_excel("~/University of Michigan Dropbox/Steven Monda/@SALAMI LAB/JAVELIN Recurrence Data/43018_2024_896_MOESM2_ESM(4).xlsx", 
           sheet = "Supp. Table 1", col_types = c("text", 
                                                  "text", "text", "text", "text", "text", 
                                                  "text", "text", "text", "numeric", 
                                                  "text", "text", "text", "numeric", 
                                                  "numeric", "text", "numeric", "numeric", 
                                                  "numeric", "numeric", "text", "numeric", 
                                                  "numeric", "numeric", "numeric", 
                                                  "numeric", "numeric", "numeric", 
                                                  "numeric", "numeric", "numeric", 
                                                  "numeric", "numeric"))
natjavclin<-subset(natjavclin,natjavclin$Cohort=="JAVELIN-101")


tpmjav<-data.frame(t(natjavrna))
colnames(tpmjav) <- natjavrna$hgnc_symbol
tpmjav<-tpmjav[-1,]
tpmjav_num <- data.frame(
  lapply(tpmjav, function(x) as.numeric(as.character(x))),
  row.names = rownames(tpmjav),
  check.names = FALSE
)
rm(natjavrna,tpmjav)

#data matching file for reconciling PMID: 39789182 with EMD Serono clinical data, with annotation of metastatic versus primary site for tumor sequencing
finalmatch <- read_excel("~/University of Michigan Dropbox/Steven Monda/@SALAMI LAB/JAVELIN Recurrence Data/finalmatchwithsitecategory.xlsx")

cancerhx$natid<-finalmatch$UUID[match(cancerhx$USUBJID,finalmatch$USUBJID)]

cancerhx$site<-finalmatch$site[match(cancerhx$USUBJID,finalmatch$USUBJID)]

cancerhx$sitespec<-finalmatch$sitspec[match(cancerhx$USUBJID,finalmatch$USUBJID)]

```

## Identifying Subcohort for Analysis - Recurrent Patients and De Novo Metastatic Patients
1 of 886 randomized patients did not have any available cancer history (e.g. prior nephrectomy) and was excluded, additional exclusions are as follows
```{r cohort identificcation}

nephreclist<-unique(subset(tte$USUBJID,tte$NEPHRCBL=="Yes"))

cancerhx$nephrec<-cancerhx$USUBJID %in% nephreclist

#initial consort diagram split - 474 and 411 patients
consort1nomet<-subset(cancerhx,!cancerhx$MHAJCIM %in% c("M1","M1A","M1B","M1C"))
consort1met<-subset(cancerhx,cancerhx$MHAJCIM %in% c("M1","M1A","M1B","M1C"))
#split for De Novo Met and Recurrent Patients, 420 recurrent patients and 296 de novo met

recurrers<-subset(consort1nomet,consort1nomet$USUBJID %in% nephreclist & consort1nomet$MHAJCIT!="T0" & !consort1nomet$MHRCTYP %in% c("NEWLY DIAGNOSED","NO RECURRENCE"))

#number of M0 patients excluded with nephrectomy
table(consort1nomet$nephrec)
#number of M0 patients  excluded not documented as recurrent
table(!consort1nomet$MHRCTYP %in% c("NEWLY DIAGNOSED","NO RECURRENCE"))
#number of M0 patients with "T0" nephrectomy
table(consort1nomet$MHAJCIT!="T0")


#identify the de novo metastatic patients
metatpresent<-subset(consort1met, consort1met$MHRCTYP %in% c("NEWLY DIAGNOSED","NO RECURRENCE","UNKNOWN"))

#number of M+ patients documented as recurrence e.g. distant recurrence, locoregional recurrence, recurrence, regional recurrence
table(consort1met$MHRCTYP %in% c("NEWLY DIAGNOSED","NO RECURRENCE","UNKNOWN"))

#final represents the final cohort for analysis
final<-rbind(recurrers,metatpresent)
final$recurrer<-ifelse(final$USUBJID %in% recurrers$USUBJID,"recur","denovo")

rm(consort1nomet,consort1met,recurrers,metatpresent)

#aligning overall survival, progression free survival, overall response rate with final clinical dataframe
fos<-subset(tte, tte$USUBJID %in% final$USUBJID & tte$PARAMCD=="OS")


final<-final[match(fos$USUBJID,final$USUBJID),]
fpfs<-subset(tte, tte$USUBJID %in% final$USUBJID & tte$PARAMCD=="PFS_P")
fpfs<-fpfs[match(final$USUBJID,fpfs$USUBJID),]
fos<-fos[match(final$USUBJID,fos$USUBJID),]
forr<-subset(response,response$USUBJID %in% final$USUBJID & response$PARAMCD=="BOR_C")
forr<-forr[match(final$USUBJID,forr$USUBJID),]

# 
fclin<-subset(clin,clin$USUBJID %in% final$USUBJID)
#time to response not lined up with final because less patients in a responder category
fttr<-subset(tte, tte$USUBJID %in% final$USUBJID & tte$PARAMCD=="TTR_C")



final$timetotrial<-ifelse(final$recurrer=="recur",final$MHTSID,NA)
final$timetotrialinc<-ifelse(final$recurrer=="recur",final$MHTSID,0)

final$timetotrialc<-ifelse(final$recurrer=="recur",
  ifelse(final$MHTSID<12,"early",
                    ifelse(
                      final$MHTSID>=12&final$MHTSID<60,"intermediate",
                      ifelse(
                        final$MHTSID>=60,"late","NAA"
                        
                      ))),NA)

final$verylate<-ifelse(final$recurrer=="recur",
                           ifelse(final$MHTSID>=120,"very late","not"
                                  
                                      
                                    ),"denovo")

final$recistlesions<-fclin$NTGSTB
final$totallesions<-fclin$NTGSTB+fclin$NNTGSTB

final$ttr<-cancerhx$ttr[match(final$USUBJID,cancerhx$USUBJID)]

final$timetorecur<-ifelse(final$recurrer=="recur",final$MHTSID-final$MHRMDUR,NA)
final$timetorecurinc<-ifelse(final$recurrer=="recur",final$MHTSID-final$MHRMDUR,0)


final$timetorecurc<-ifelse(final$timetorecur<12,"early",
                    ifelse(
                      final$timetorecur>=12&final$timetorecur<60,"intermediate",
                      ifelse(
                        final$timetorecur>=60,"late","NAA"
                        
                      )))

#reconcile available clinical variables from 
final$sarc<-ifelse(final$natid %in% subset(natjavclin$UUID,natjavclin$Sarcomatoid=="Yes"),TRUE,
                   ifelse(final$natid %in% subset(natjavclin$UUID,natjavclin$Sarcomatoid=="No"),"FALSE",NA))

final$sitespecclean<-ifelse(final$sitespec %in% c("BRAIN","COLON","SPLEEN"),"OTHER",final$sitespec)
final$wgii<-natjavclin$wGII[match(final$natid,natjavclin$UUID)]
final$fscounts<-natjavclin$FS_Counts[match(final$natid,natjavclin$UUID)]
final$tmb<-natjavclin$TMB_Counts[match(final$natid,natjavclin$UUID)]

final$shrink<-as.numeric(natjavclin$Tumor_Shrinkage[match(final$natid,natjavclin$UUID)])

#Exceptional Response Paper for Mutational Results

natjavmut<-read_excel("~/Library/CloudStorage/OneDrive-MichiganMedicine/Research/M vs P/RDS121_2024_12_12T10_04_unzip/43018_2024_896_MOESM2_ESM(4).xlsx", 
                      sheet = "Supp. Table 4")
natjavmut<-subset(natjavmut,natjavmut$UUID %in% natjavclin$UUID)

natjavmut$type<-ifelse(natjavmut$Variant_Classification=="Silent",NA,
ifelse(natjavmut$Variant_Classification=="Nonsense_Mutation","Nonsense",
   ifelse(natjavmut$Variant_Classification=="Missense_Mutation","Missense", 
             ifelse(natjavmut$Variant_Classification %in% c("Frame_Shift_Del","Frame_Shift_Ins"),"Frameshift",
ifelse(natjavmut$Variant_Classification %in% c("In_Frame_Del","In_Frame_Ins"),"Inframe","Other")))))
                    
  


final$tls<-natjavclin$TLS_Meylan[match(final$natid,natjavclin$UUID)]
final$ith<-as.numeric(natjavclin$ITH[match(final$natid,natjavclin$UUID)])
final$ith[is.infinite(final$ith)] <- NA

final$purity<-natjavclin$Purity[match(final$natid,natjavclin$UUID)]
final$neoant<-natjavclin$Clonal_Neoantigen_Load[match(final$natid,natjavclin$UUID)]
final$ith<-natjavclin$ITH[match(final$natid,natjavclin$UUID)]
final$clonalcall<-final$natid %in% subset(natjavmut$UUID,!is.na(natjavmut$mutation_clonality))
final$ploidy<-natjavclin$Ploidy[match(final$natid,natjavclin$UUID)]


#imdc corrected is used throughout for all downstream analyses, this excludes the time to diagnosis variable as this is embedded within recurrence time so this is an attempt to eliminate structural colinearity, how imdc is represented in the model is assessed further below
final$imdccorrected<-final$HENGBLN-(final$MHTSID<12)

final$grade<-ifelse(final$MHHISG==""&final$MHFUHR=="",NA,ifelse(
  final$MHHISG!="",final$MHHISG,final$MHFUHR))

final$tstage<-ifelse(final$MHAJCIT %in% c("T1","T1A","T1B"),"T1",
                     ifelse(final$MHAJCIT %in% c("T2","T2A","T2B"),"T2", 
                            ifelse(final$MHAJCIT %in% c("T3","T3A","T3B","T3C","T4"),"T3+",
                                   ifelse(final$MHAJCIT %in% c("TX","T0",""),NA,NA 
                                   ))))
final$initialnstage<-ifelse(final$MHAJCIN %in% c("N1","N2","N3"),"N+",
                     ifelse(final$MHAJCIN %in% c("NX","T2A","T2B"),"Nx", 
                            ifelse(final$MHAJCIN %in% c("N0","T3A","T3B","T3C","T4"),"N0",NA
                                 
                                   )))

final$categories<-ifelse(final$recurrer!="recur","De Novo Metastatic",final$timetotrialc)

```
## Stacked Bar Plot for Figure 1
```{r figure 1}


final$tstage2 <- ifelse(is.na(final$tstage), "Unknown", as.character(final$tstage))

final$tstage2 <- factor(final$tstage2,
                      levels = c("Unknown","T1","T2","T3+"))

final$grade2 <- ifelse(is.na(final$grade), "Unknown", gsub("GRADE ", "", final$grade))
final$grade2 <- factor(final$grade2,
                      levels = c("Unknown", "1", "2", "3", "4"))

final$sarc2 <- ifelse(is.na(final$sarc), "Unknown", tools::toTitleCase(tolower(as.character(final$sarc))))
final$sarc2 <- factor(final$sarc2,
                      levels = c("Unknown", "False", "True"))

final$imdccorrected2 <- ifelse(is.na(final$imdccorrected), "Unknown",ifelse(final$imdccorrected>2, ">2", as.character(final$imdccorrected)))
final$imdccorrected2 <- factor(final$imdccorrected2,
                      levels = c("Unknown","0","1","2"
                                 ,">2"))

pal_tstage <- c("Unknown"="gray85", "T1"="#b8e186", "T2"="#fde0ef", "T3+"="#c51b7d")

pal_grade <- c("Unknown"="gray85",
               "1"="#74add1",
               "2"="#abd9e9",
               "3"="#fdae61",
               "4"="#d73027")

pal_sarc <- c("Unknown"="gray85", "False"="#e0f3f8", "True"="#a50026")

pal_imdc <- c("Unknown"="gray85", "0"="#1a9850", "1"="#a6d96a", "2"="#fee08b", ">2"="#f46d43")

sub<-subset(final,final$categories=="De Novo Metastatic")

p1 <- ggplot(sub, aes(x = 1, fill = tstage2)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = pal_tstage, drop = FALSE, name = "T-stage")+
  labs(y = NULL, x = "T-stage") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        axis.title.y = element_text(angle = 90, vjust = 0.5))


p2 <- ggplot(sub, aes(x = 1, fill = grade2)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) + 
  scale_fill_manual(values = pal_grade, drop = FALSE, name = "Grade")+
  labs(y = NULL, x = "Grade") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
                axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        axis.title.x = element_text(angle = 90, vjust = 0.5))

p3 <- ggplot(sub, aes(x = 1, fill = factor(imdccorrected2))) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = pal_imdc, drop = FALSE, name = "Adjusted IMDC")+
  labs(y = NULL, x = "IMDC*") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
                axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        axis.title.x = element_text(angle = 90, vjust = 0.5))

p4 <- ggplot(sub, aes(x = 1, fill = factor(sarc2))) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = pal_sarc, drop = FALSE, name = "Sarcomatoid")+
  labs(y = NULL, x = "Sarc.") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank(),
        axis.title.x = element_text(angle = 90, vjust = 0.5))

leg_tstage <- get_legend(
  p1 + theme(legend.position="bottom")
)

leg_grade <- get_legend(
  p2 +
    theme(legend.position = "bottom") +
    guides(fill = guide_legend(nrow = 1, byrow = TRUE))
)

leg_imdc <- get_legend(
  p3 + theme(legend.position="bottom")
)

leg_sarc <- get_legend(
  p4 + theme(legend.position="bottom")
)
leg_all <- plot_grid(
  leg_tstage,
  leg_grade,
  leg_imdc,
  leg_sarc,
  nrow = 2
)


  p1 <- p1 + theme(legend.position="none",
     plot.margin = ggplot2::margin(0, 0, 0, 0))
  p2 <- p2 + theme(legend.position="none",
    plot.margin = ggplot2::margin(0, 0, 0, 0))
  p3 <- p3 + theme(legend.position="none",
  plot.margin = ggplot2::margin(0, 0, 0, 0))
  p4 <- p4 + theme(legend.position="none",
   plot.margin = ggplot2::margin(0, 0, 0, 0))
  
p_combined <- plot_grid(
  p1, p2, p3,p4,
  ncol = 1,
  align = "v",
  axis = "lr",
  rel_heights = c(1, 1, 1)
)

p_denovo<-p_combined


sub<-subset(final,final$categories=="early")

p1 <- ggplot(sub, aes(x = 1, fill = tstage2)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = pal_tstage, drop = FALSE, name = "T-stage")+
  labs(y = NULL, x = "T-stage") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        axis.title.y = element_text(angle = 90, vjust = 0.5))


p2 <- ggplot(sub, aes(x = 1, fill = grade2)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) + 
  scale_fill_manual(values = pal_grade, drop = FALSE, name = "Grade")+
  labs(y = NULL, x = "Grade") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
                axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        axis.title.x = element_text(angle = 90, vjust = 0.5))

p3 <- ggplot(sub, aes(x = 1, fill = factor(imdccorrected2))) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = pal_imdc, drop = FALSE, name = "Adjusted IMDC")+
  labs(y = NULL, x = "IMDC*") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
                axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        axis.title.x = element_text(angle = 90, vjust = 0.5))

p4 <- ggplot(sub, aes(x = 1, fill = factor(sarc2))) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = pal_sarc, drop = FALSE, name = "Sarcomatoid")+
  labs(y = NULL, x = "Sarc.") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank(),
        axis.title.x = element_text(angle = 90, vjust = 0.5))

  p1 <- p1 + theme(legend.position="none",
     plot.margin = ggplot2::margin(0, 0, 0, 0))
  p2 <- p2 + theme(legend.position="none",
    plot.margin = ggplot2::margin(0, 0, 0, 0))
  p3 <- p3 + theme(legend.position="none",
  plot.margin = ggplot2::margin(0, 0, 0, 0))
  p4 <- p4 + theme(legend.position="none",
   plot.margin = ggplot2::margin(0, 0, 0, 0))

p_combined <- plot_grid(
  p1, p2, p3,p4,
  ncol = 1,
  align = "v",
  axis = "lr",
  rel_heights = c(1, 1, 1)
)

p_early<-p_combined

sub<-subset(final,final$categories=="intermediate")

p1 <- ggplot(sub, aes(x = 1, fill = tstage2)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = pal_tstage, drop = FALSE, name = "T-stage")+
  labs(y = NULL, x = "T-stage") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        axis.title.y = element_text(angle = 90, vjust = 0.5))


p2 <- ggplot(sub, aes(x = 1, fill = grade2)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) + 
  scale_fill_manual(values = pal_grade, drop = FALSE, name = "Grade")+
  labs(y = NULL, x = "Grade") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
                axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        axis.title.x = element_text(angle = 90, vjust = 0.5))

p3 <- ggplot(sub, aes(x = 1, fill = factor(imdccorrected2))) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = pal_imdc, drop = FALSE, name = "Adjusted IMDC")+
  labs(y = NULL, x = "IMDC*") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
                axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        axis.title.x = element_text(angle = 90, vjust = 0.5))

p4 <- ggplot(sub, aes(x = 1, fill = factor(sarc2))) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = pal_sarc, drop = FALSE, name = "Sarcomatoid")+
  labs(y = NULL, x = "Sarc.") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank(),
        axis.title.x = element_text(angle = 90, vjust = 0.5))

  p1 <- p1 + theme(legend.position="none",
     plot.margin = ggplot2::margin(0, 0, 0, 0))
  p2 <- p2 + theme(legend.position="none",
    plot.margin = ggplot2::margin(0, 0, 0, 0))
  p3 <- p3 + theme(legend.position="none",
  plot.margin = ggplot2::margin(0, 0, 0, 0))
  p4 <- p4 + theme(legend.position="none",
   plot.margin = ggplot2::margin(0, 0, 0, 0))

p_combined <- plot_grid(
  p1, p2, p3,p4,
  ncol = 1,
  align = "v",
  axis = "lr",
  rel_heights = c(1, 1, 1)
)

p_int<-p_combined


sub<-subset(final,final$categories=="late")

p1 <- ggplot(sub, aes(x = 1, fill = tstage2)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = pal_tstage, drop = FALSE, name = "T-stage")+
  labs(y = NULL, x = "T-stage") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        axis.title.y = element_text(angle = 90, vjust = 0.5))


p2 <- ggplot(sub, aes(x = 1, fill = grade2)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) + 
  scale_fill_manual(values = pal_grade, drop = FALSE, name = "Grade")+
  labs(y = NULL, x = "Grade") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
                axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        axis.title.x = element_text(angle = 90, vjust = 0.5))

p3 <- ggplot(sub, aes(x = 1, fill = factor(imdccorrected2))) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = pal_imdc, drop = FALSE, name = "Adjusted IMDC")+
  labs(y = NULL, x = "IMDC*") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
                axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        axis.title.x = element_text(angle = 90, vjust = 0.5))

p4 <- ggplot(sub, aes(x = 1, fill = factor(sarc2))) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = pal_sarc, drop = FALSE, name = "Sarcomatoid")+
  labs(y = NULL, x = "Sarc.") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank(),
        axis.title.x = element_text(angle = 90, vjust = 0.5))

  p1 <- p1 + theme(legend.position="none",
     plot.margin = ggplot2::margin(0, 0, 0, 0))
  p2 <- p2 + theme(legend.position="none",
    plot.margin = ggplot2::margin(0, 0, 0, 0))
  p3 <- p3 + theme(legend.position="none",
  plot.margin = ggplot2::margin(0, 0, 0, 0))
  p4 <- p4 + theme(legend.position="none",
   plot.margin = ggplot2::margin(0, 0, 0, 0))

p_combined <- plot_grid(
  p1, p2, p3,p4,
  ncol = 1,
  align = "v",
  axis = "lr",
  rel_heights = c(1, 1, 1)
)

p_late<-p_combined

p_top <- plot_grid(
  p_denovo, p_early,p_int,p_late,
  ncol = 4,
  align = "h"
)

leg_all <- plot_grid(
  leg_tstage,
  leg_grade,
  leg_imdc,
  leg_sarc,
  nrow = 2
)

margin_theme <- theme(plot.margin = ggplot2::margin(t = 5, r = 5, b = 15, l = 5))

p_top <- plot_grid(
  p_denovo + margin_theme,
  p_early + margin_theme,
  p_int + margin_theme,
  p_late + margin_theme,
  ncol = 4,
  align = "h")

  final_plot <- plot_grid(
  p_top,
  leg_all,
  ncol = 1,
  rel_heights = c(1, 0.2)
)



final_plot
#to be combined with biorender

```




## Establishing Data Sets for Survival Analysis e.g. Prognostic Association of Recurrence Status and 

```{r dataset housekeeping for analysis}

fpfs$categories<-final$categories
fos$categories<-final$categories
fpfs$imdc<-final$imdccorrected

fpfs$timetotrial<-final$timetotrial
fpfs$timetotrialinc<-final$timetotrialinc

fpfs$heng<-final$HENGBL
fos$imdc<-final$imdccorrected
fos$heng<-final$HENGBL

final$ecog<-fos$STRATI1N




```


##Table 1 Clinical Data Summary

```{r clinical data Summary}


final$pdl1<-ifelse(final$pdl1=="Y",TRUE,
                   ifelse(final$pdl1=="N",FALSE,NA))
                   
final$age<-as.numeric(final$age)
tbl_summary(
  final,
  by = categories,
  include = c(
    SEX, age, ecog,nephrec, MHLAT, tstage,initialnstage, grade,
    pdl1, imdccorrected, sarc, site
  ),
  type = list(age ~ "continuous"),
  statistic = list(age ~ "{median} ({p25}, {p75})")  # optional formatting
)



```

## Comparing aggressive clinical features
```{r Aggressive clinical feature comparison}



tab <- table(final$categories, final$tstage=="T3+")

binom.confint(
    tab[, "TRUE"],
    rowSums(
      tab),
    method = "wilson"
)

anova(glm(final$tstage=="T3+"~ final$categories, family = binomial),test="LRT")



tab <- table(final$categories, final$initialnstage == "N+")

binom.confint(
    tab[, "TRUE"],
    rowSums(
      tab),
    method = "wilson"
)

anova(glm(final$initialnstage == "N+"~ final$categories, family = binomial),test="LRT")




tab <- table(final$categories, final$grade %in% c("GRADE 3","GRADE 4"))

binom.confint(
    tab[, "TRUE"],
    rowSums(
      tab),
    method = "wilson"
)

tab <- table(final$categories, final$sarc=="TRUE")

binom.confint(
    tab[, "TRUE"],
    rowSums(
      tab),
    method = "wilson"
)

anova(glm(final$sarc=="TRUE" ~ final$categories, family = binomial),test="LRT")



tab <- table(final$categories, final$pdl1)

binom.confint(
    tab[, "TRUE"],
    rowSums(
      tab),
    method = "wilson"
)

anova(glm(final$pdl1=="TRUE" ~ final$categories, family = binomial),test="LRT")



tab <- table(final$categories, final$imdccorrected>1)

binom.confint(
    tab[, "TRUE"],
    rowSums(
      tab),
    method = "wilson"
)

anova(glm(final$imdccorrected>1 ~ final$categories, family = binomial),test="LRT")

```


## Mutation Analysis

```{r mutation analysis}

  #479 in mutation dataframe, 479 with genes included
#for this VHL, PBRM1, BAP1, SETD2, KDM5C, MTOR, TSC2, PIK3CA, PTEN, TP53



#all patients with TMB are listed in mutation file
#register cna level data

CNAs <- read_delim("~/Library/CloudStorage/OneDrive-MichiganMedicine/Research/recurrence javelin/gistic_CNAs_JAVELIN101_UUID.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)

final$cna9p213<-CNAs$`9p21.3`[match(final$natid,CNAs$sample)]
final$cna3p2131<-CNAs$`3p21.31`[match(final$natid,CNAs$sample)]
final$cna14q112<-CNAs$`14q11.2`[match(final$natid,CNAs$sample)]
final$cna14q3233<-CNAs$`14q32.33`[match(final$natid,CNAs$sample)]
final$cna5q351<-CNAs$`5q35.1`[match(final$natid,CNAs$sample)]
final$cna5q313<-CNAs$`5q31.3`[match(final$natid,CNAs$sample)]


#?TP53?
final$cna17p131<-CNAs$`17p13.1`[match(final$natid,CNAs$sample)]

final$cna17q253<-CNAs$`17q25.3`[match(final$natid,CNAs$sample)]
#MYC Gain- not that impressive
final$cna8q243<-CNAs$`8q24.3`[match(final$natid,CNAs$sample)]
final$cna1p3611<-CNAs$`1p36.11`[match(final$natid,CNAs$sample)]
final$cna1p3631<-CNAs$`1p36.31`[match(final$natid,CNAs$sample)]
final$cna8q243<-CNAs$`8q24.3`[match(final$natid,CNAs$sample)]
final$cna20q13<-CNAs$`20q13.33`[match(final$natid,CNAs$sample)]



fit <- lm(final$cna9p213=="-1" ~ categories + site, data = final)
# Reduced model (no recurrence timing)
fit_reduced <- lm(cna9p213=="-1" ~ site, data = final)

final$loss9p<- final$cna9p213 == "-1"
final$loss3p <- final$cna3p2131 == "-1"
final$loss14q <- final$cna14q3233 == "-1"

final$ch9p<- ifelse(final$cna9p213 == "-1","Deletion","")
final$ch3p <- ifelse(final$cna3p2131 == "-1","Deletion","")
final$ch14q <- ifelse(final$cna14q3233 == "-1","Deletion","")
# 
# fit_reduced <- glm(loss9p ~ site, data = final, family = binomial)
# fit         <- glm(loss9p ~ categories + site , data = final, family = binomial)
# anova(fit_reduced, fit, test= "LRT")
# 
# fit_reduced <- glm(loss3p ~ site, data = final, family = binomial)
# fit         <- glm(loss3p ~ categories + site , data = final, family = binomial)
# anova(fit_reduced, fit, test= "LRT")
# 
# fit_reduced <- glm(loss14q ~ site, data = final, family = binomial)
# fit         <- glm(loss14q ~ categories + site , data = final, family = binomial)
# anova(fit_reduced, fit, test= "LRT")
# 

# Likelihood ratio test




final$CNA<-final$natid %in% CNAs$sample
final$WES<-final$natid %in% natjavmut$UUID
final$VHL<-ifelse(!final$WES,NA,
       subset(natjavmut$type,natjavmut$Hugo_Symbol=="VHL"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="VHL"&!is.na(natjavmut$type)))] )
final$VHL<-ifelse(!is.na(final$VHL),final$VHL,
                  ifelse(final$WES,"",NA
                  ))
final$VHLcl<-ifelse(!final$WES,NA,
       subset(natjavmut$mutation_clonality,natjavmut$Hugo_Symbol=="VHL"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="VHL"&!is.na(natjavmut$type)))] )     

final$VHLcl<-ifelse(!is.na(final$VHLcl),final$VHLcl,
                     ifelse(final$clonalcall,"",NA))
        

#----
final$PBRM1<-ifelse(!final$WES,NA,
       subset(natjavmut$type,natjavmut$Hugo_Symbol=="PBRM1"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="PBRM1"&!is.na(natjavmut$type)))])
final$PBRM1<-ifelse(!is.na(final$PBRM1),final$PBRM1,
                  ifelse(final$WES,"",NA
                  ))
         
final$PBRM1cl<-ifelse(!final$WES,NA,
       subset(natjavmut$mutation_clonality,natjavmut$Hugo_Symbol=="PBRM1"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="PBRM1"&!is.na(natjavmut$type)))])


final$PBRM1cl<-ifelse(!is.na(final$PBRM1cl),final$PBRM1cl,
                     ifelse(final$clonalcall,"",NA))
        
#----
final$BAP1<-ifelse(!final$WES,NA,
       subset(natjavmut$type,natjavmut$Hugo_Symbol=="BAP1"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="BAP1"&!is.na(natjavmut$type)))])
final$BAP1<-ifelse(!is.na(final$BAP1),final$BAP1,
                  ifelse(final$WES,"",NA
                  ))

final$BAP1cl<-ifelse(!final$WES,NA,
       subset(natjavmut$mutation_clonality,natjavmut$Hugo_Symbol=="BAP1"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="BAP1"&!is.na(natjavmut$type)))])


final$BAP1cl<-ifelse(!is.na(final$BAP1cl),final$BAP1cl,
                     ifelse(final$clonalcall,"",NA))

#----
final$SETD2<-ifelse(!final$WES,NA,
       subset(natjavmut$type,natjavmut$Hugo_Symbol=="SETD2"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="SETD2"&!is.na(natjavmut$type)))] )
final$SETD2<-ifelse(!is.na(final$SETD2),final$SETD2,
                  ifelse(final$WES,"",NA
                  ))

final$SETD2cl<-ifelse(!final$WES,NA,
       subset(natjavmut$mutation_clonality,natjavmut$Hugo_Symbol=="SETD2"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="SETD2"&!is.na(natjavmut$type)))] )


final$SETD2cl<-ifelse(!is.na(final$SETD2cl),final$SETD2cl,
                     ifelse(final$clonalcall,"",NA))

#----
final$KDM5C<-ifelse(!final$WES,NA,
       subset(natjavmut$type,natjavmut$Hugo_Symbol=="KDM5C"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="KDM5C"&!is.na(natjavmut$type)))])


final$KDM5C<-ifelse(!is.na(final$KDM5C),final$KDM5C,
                  ifelse(final$WES,"",NA
                  ))

final$KDM5Ccl<-ifelse(!final$WES,NA,
       subset(natjavmut$mutation_clonality,natjavmut$Hugo_Symbol=="KDM5C"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="KDM5C"&!is.na(natjavmut$type)))]
                        )


final$KDM5Ccl<-ifelse(!final$WES,NA,
       subset(natjavmut$mutation_clonality,natjavmut$Hugo_Symbol=="KDM5C"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="KDM5C"&!is.na(natjavmut$type)))])

final$KDM5Ccl<-ifelse(!is.na(final$KDM5Ccl),final$KDM5Ccl,
                     ifelse(final$clonalcall,"",NA))

#----
final$MTOR<-ifelse(!final$WES,NA,
       subset(natjavmut$type,natjavmut$Hugo_Symbol=="MTOR"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="MTOR"&!is.na(natjavmut$type)))])

final$MTORcl<-ifelse(!final$WES,NA,
       subset(natjavmut$mutation_clonality,natjavmut$Hugo_Symbol=="MTOR"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="MTOR"&!is.na(natjavmut$type)))]
                        )

final$MTORcl<-ifelse(!is.na(final$MTORcl),final$MTORcl,
                     ifelse(final$clonalcall,"",NA))

final$MTOR<-ifelse(!is.na(final$MTOR),final$MTOR,
                  ifelse(final$WES,"",NA
                  ))

#----
final$TSC2<-ifelse(!final$WES,NA,
       subset(natjavmut$type,natjavmut$Hugo_Symbol=="TSC2"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="TSC2"&!is.na(natjavmut$type)))])
final$TSC2<-ifelse(!is.na(final$TSC2),final$TSC2,
                  ifelse(final$WES,"",NA
                  ))
final$TSC2cl<-ifelse(!final$WES,NA,
       subset(natjavmut$mutation_clonality,natjavmut$Hugo_Symbol=="TSC2"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="TSC2"&!is.na(natjavmut$type)))])

final$TSC2cl<-ifelse(!is.na(final$TSC2cl),final$TSC2cl,
                     ifelse(final$clonalcall,"",NA))

#----
final$PIK3CA<-ifelse(!final$WES,NA,
       subset(natjavmut$type,natjavmut$Hugo_Symbol=="PIK3CA"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="PIK3CA"&!is.na(natjavmut$type)))])

final$PIK3CAcl<-ifelse(!final$WES,NA,
       subset(natjavmut$mutation_clonality,natjavmut$Hugo_Symbol=="PIK3CA"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="PIK3CA"&!is.na(natjavmut$type)))])

final$PIK3CAcl<-ifelse(!is.na(final$PIK3CAcl),final$PIK3CAcl,
                     ifelse(final$clonalcall,"",NA))

final$PIK3CA<-ifelse(!is.na(final$PIK3CA),final$PIK3CA,
                  ifelse(final$WES,"",NA
                  ))

#----
final$TP53<-ifelse(!final$WES,NA,
       subset(natjavmut$type,natjavmut$Hugo_Symbol=="TP53"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="TP53"&!is.na(natjavmut$type)))])

final$TP53cl<-ifelse(!final$WES,NA,
       subset(natjavmut$mutation_clonality,natjavmut$Hugo_Symbol=="TP53"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="TP53"&!is.na(natjavmut$type)))])

final$TP53cl<-ifelse(!is.na(final$TP53cl),final$TP53cl,
                     ifelse(final$clonalcall,"",NA))

final$TP53<-ifelse(!is.na(final$TP53),final$TP53,
                  ifelse(final$WES,"",NA
                  ))

#----
final$PTEN<-ifelse(!final$WES,NA,
       subset(natjavmut$type,natjavmut$Hugo_Symbol=="PTEN"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="PTEN"&!is.na(natjavmut$type)))])

final$PTENcl<-ifelse(!final$WES,NA,
       subset(natjavmut$mutation_clonality,natjavmut$Hugo_Symbol=="PTEN"&!is.na(natjavmut$type))
       [match(final$natid,                    subset(natjavmut$UUID,natjavmut$Hugo_Symbol=="PTEN"&!is.na(natjavmut$type)))])

final$PTENcl<-ifelse(!is.na(final$PTENcl),final$PTENcl,
                     ifelse(final$clonalcall,"",NA))

final$PTEN<-ifelse(!is.na(final$PTEN),final$PTEN,                  ifelse(final$WES,"",NA
                  ))

#479 in mutation dataframe, 479 with genes included
#for this VHL, PBRM1, BAP1, SETD2, KDM5C, MTOR, TSC2, PIK3CA, PTEN, TP53
oncoprint<-subset(final,final$WES&final$categories=="De Novo Metastatic")

genes <- c(
  "VHL","PBRM1","BAP1","SETD2","KDM5C",
  "MTOR","PIK3CA","PTEN","TP53"
)

mat <- t(as.matrix(oncoprint[, genes]))
colnames(mat) <- oncoprint$natid


col <- c(
  Missense = "#e6ab02",
  Nonsense = "#d95f02",
  Frameshift = "#66a61e",
  Inframe = "#e7298a",
  Other = "#7570b3"
)


alter_fun <- list(
  background = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = "#F0F0F0", col = NA)
    ),

  Missense = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Missense"], col = NA)
    ),

  Nonsense = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Nonsense"], col = NA)
    ),

  Frameshift = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Frameshift"], col = NA)
    ),

  Inframe = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Inframe"], col = NA)
    ),

  Other = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Other"], col = NA)
    )
)

wgii_col <- colorRamp2(
  c(min(oncoprint$wgii, na.rm = TRUE),
    median(oncoprint$wgii, na.rm = TRUE),
    max(oncoprint$wgii, na.rm = TRUE)),
  c("#F0F0F0", "#feb24c", "#bd0026")
)

oncoprint$logTMB <- log10(oncoprint$tmb + 1)

tmb_col <- colorRamp2(
  c(min(oncoprint$logTMB, na.rm = TRUE),
    median(oncoprint$logTMB, na.rm = TRUE),
    max(oncoprint$logTMB, na.rm = TRUE)),
  c("#F7FBFF", "#9ecae1", "#08306B")
)

top_anno <- HeatmapAnnotation(
  WGII = oncoprint$wgii,
  TMB = oncoprint$logTMB,
  col = list(
    WGII = wgii_col,
    TMB = tmb_col
  ),
  annotation_name_side = "left",
  simple_anno_size = unit(7, "mm")
)


dnonco <- oncoPrint(
  mat,
  alter_fun = alter_fun,
  col = col,
  top_annotation = top_anno,
  remove_empty_columns = TRUE,
  remove_empty_rows = TRUE,
  show_column_names = FALSE,
  show_row_names = TRUE,
  row_names_side = "left",
  row_order = seq_len(nrow(mat)),
  pct_side="right",
  right_annotation = NULL
)
dn<-draw(dnonco, show_heatmap_legend = FALSE,
  show_annotation_legend = FALSE,)




##########
oncoprint<-subset(final,final$WES
                  &final$categories=="early"
                  )

genes <- c(
  "VHL","PBRM1","BAP1","SETD2","KDM5C",
  "MTOR","PIK3CA","PTEN","TP53"
)

mat <- t(as.matrix(oncoprint[, genes]))
colnames(mat) <- oncoprint$natid


col <- c(
  Missense = "#e6ab02",
  Nonsense = "#d95f02",
  Frameshift = "#66a61e",
  Inframe = "#e7298a",
  Other = "#7570b3"
)


alter_fun <- list(
  background = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = "#F0F0F0", col = NA)
    ),

  Missense = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Missense"], col = NA)
    ),

  Nonsense = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Nonsense"], col = NA)
    ),

  Frameshift = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Frameshift"], col = NA)
    ),

  Inframe = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Inframe"], col = NA)
    ),

  Other = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Other"], col = NA)
    )
)

wgii_col <- colorRamp2(
  c(min(oncoprint$wgii, na.rm = TRUE),
    median(oncoprint$wgii, na.rm = TRUE),
    max(oncoprint$wgii, na.rm = TRUE)),
  c("#F0F0F0", "#feb24c", "#bd0026")
)

oncoprint$logTMB <- log10(oncoprint$tmb + 1)

tmb_col <- colorRamp2(
  c(min(oncoprint$logTMB, na.rm = TRUE),
    median(oncoprint$logTMB, na.rm = TRUE),
    max(oncoprint$logTMB, na.rm = TRUE)),
  c("#F7FBFF", "#9ecae1", "#08306B")
)

top_anno <- HeatmapAnnotation(
  WGII = oncoprint$wgii,
  TMB = oncoprint$logTMB,
  col = list(
    WGII = wgii_col,
    TMB = tmb_col
  ),
show_annotation_name = FALSE, 
  simple_anno_size = unit(7, "mm")
)


eonco <- oncoPrint(
  mat,
  alter_fun = alter_fun,
  col = col,
  top_annotation = top_anno,
  remove_empty_columns = TRUE,
  remove_empty_rows = TRUE,
  show_column_names = FALSE,
  show_row_names = FALSE,
  row_names_side = "left",
  row_order = seq_len(nrow(mat)),
  right_annotation = NULL,
  pct_side="right"
)
e<-draw(eonco, show_heatmap_legend = FALSE,
  show_annotation_legend = FALSE,)

##########
oncoprint<-subset(final,final$WES
                  &final$categories=="intermediate"
                  )

genes <- c(
  "VHL","PBRM1","BAP1","SETD2","KDM5C",
  "MTOR","PIK3CA","PTEN","TP53"
)

mat <- t(as.matrix(oncoprint[, genes]))
colnames(mat) <- oncoprint$natid


col <- c(
  Missense = "#e6ab02",
  Nonsense = "#d95f02",
  Frameshift = "#66a61e",
  Inframe = "#e7298a",
  Other = "#7570b3"
)


alter_fun <- list(
  background = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = "#F0F0F0", col = NA)
    ),

  Missense = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Missense"], col = NA)
    ),

  Nonsense = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Nonsense"], col = NA)
    ),

  Frameshift = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Frameshift"], col = NA)
    ),

  Inframe = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Inframe"], col = NA)
    ),

  Other = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Other"], col = NA)
    )
)

wgii_col <- colorRamp2(
  c(min(oncoprint$wgii, na.rm = TRUE),
    median(oncoprint$wgii, na.rm = TRUE),
    max(oncoprint$wgii, na.rm = TRUE)),
  c("#F0F0F0", "#feb24c", "#bd0026")
)

oncoprint$logTMB <- log10(oncoprint$tmb + 1)

tmb_col <- colorRamp2(
  c(min(oncoprint$logTMB, na.rm = TRUE),
    median(oncoprint$logTMB, na.rm = TRUE),
    max(oncoprint$logTMB, na.rm = TRUE)),
  c("#F7FBFF", "#9ecae1", "#08306B")
)

top_anno <- HeatmapAnnotation(
  WGII = oncoprint$wgii,
  TMB = oncoprint$logTMB,
  col = list(
    WGII = wgii_col,
    TMB = tmb_col
  ),show_annotation_name = FALSE, 
  simple_anno_size = unit(7, "mm")
)


ionco <- oncoPrint(
  mat,
  alter_fun = alter_fun,
  col = col,
  top_annotation = top_anno,
  remove_empty_columns = TRUE,
  remove_empty_rows = TRUE,
  show_column_names = FALSE,
  show_row_names = FALSE,
  row_names_side = "left",
  row_order = seq_len(nrow(mat)),
  right_annotation = NULL,
  pct_side="right"
)
i<-draw(ionco, show_heatmap_legend = FALSE,
  show_annotation_legend = FALSE,)

finalm<-subset(final,final$site=="KIDNEY")

##########
oncoprint<-subset(final,final$WES
                  &final$categories=="late"
                  )

genes <- c(
  "VHL","PBRM1","BAP1","SETD2","KDM5C",
  "MTOR","PIK3CA","PTEN","TP53"
)

mat <- t(as.matrix(oncoprint[, genes]))
colnames(mat) <- oncoprint$natid


col <- c(
  Missense = "#e6ab02",
  Nonsense = "#d95f02",
  Frameshift = "#66a61e",
  Inframe = "#e7298a",
  Other = "#7570b3"
)


alter_fun <- list(
  background = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = "#F0F0F0", col = NA)
    ),

  Missense = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Missense"], col = NA)
    ),

  Nonsense = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Nonsense"], col = NA)
    ),

  Frameshift = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Frameshift"], col = NA)
    ),

  Inframe = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Inframe"], col = NA)
    ),

  Other = function(x, y, w, h)
    grid::grid.rect(x, y, w, h,
      gp = grid::gpar(fill = col["Other"], col = NA)
    )
)

wgii_col <- colorRamp2(
  c(min(oncoprint$wgii, na.rm = TRUE),
    median(oncoprint$wgii, na.rm = TRUE),
    max(oncoprint$wgii, na.rm = TRUE)),
  c("#F0F0F0", "#feb24c", "#bd0026")
)

oncoprint$logTMB <- log10(oncoprint$tmb + 1)

tmb_col <- colorRamp2(
  c(min(oncoprint$logTMB, na.rm = TRUE),
    median(oncoprint$logTMB, na.rm = TRUE),
    max(oncoprint$logTMB, na.rm = TRUE)),
  c("#F7FBFF", "#9ecae1", "#08306B")
)

top_anno <- HeatmapAnnotation(
  WGII = oncoprint$wgii,
  TMB = oncoprint$logTMB,
  col = list(
    WGII = wgii_col,
    TMB = tmb_col
  ),
show_annotation_name = FALSE, 
  simple_anno_size = unit(7, "mm")
)


lonco <- oncoPrint(
  mat,
  alter_fun = alter_fun,
  col = col,
  top_annotation = top_anno,
  remove_empty_columns = TRUE,
  remove_empty_rows = TRUE,
  show_column_names = FALSE,
  show_row_names = FALSE,
  row_names_side = "left",
  row_order = seq_len(nrow(mat)),
  right_annotation = NULL,
  
  pct_side="right"
)
l<-draw(
  lonco,
  heatmap_legend_side = "right",
  annotation_legend_side = "right",
  merge_legends = TRUE
)

ht <- draw(lonco)

dntitle <- plot_grid( ggdraw() +
    draw_label("De Novo Metastatic", fontface = "bold", size = 12), grid.grabExpr(draw(dn)), ncol=1,rel_heights = c(.1,1)
)

etitle <- plot_grid( ggdraw() +
    draw_label("Early Recurrent      ", fontface = "bold", size = 12), grid.grabExpr(draw(e)), ncol=1,rel_heights = c(.1,1)
)

ititle <- plot_grid( ggdraw() +
    draw_label("Intermediate Recurrent   ", fontface = "bold", size = 12), grid.grabExpr(draw(i)), ncol=1,rel_heights = c(.1,1)
)

ltitle <- plot_grid( ggdraw() +
    draw_label("Late Recurrent                           ", fontface = "bold", size = 12), grid.grabExpr(draw(l)), ncol=1,rel_heights = c(.1,1)
)

combonco<-plot_grid(
dntitle,etitle,ititle,ltitle,
  ncol = 4,rel_widths = c(1.3,1,1,1.5)
)

```

#WGII figure and TMB analysis 
```{r wgii as continuous variable}
#wgii

# modeling variable
final$categories <- factor(
  final$categories,
  levels = c("De Novo Metastatic", "early", "intermediate", "late")
)

final$categories_model <- relevel(final$categories, ref = "late")

# plotting order
final$categories_plot <- factor(
  final$categories_model,
  levels = c("De Novo Metastatic","early","intermediate","late")
)

fit <- lm(wgii ~ categories_model + site, data = final)
final$predwgiicat<-predict(fit,newdata = final)
summary(fit)
cbind(
  coef(fit),
  confint(fit)
)
# Reduced model (no recurrence timing)
fit_reduced <- lm(wgii ~ site, data = final)

# Likelihood ratio test
anova(fit_reduced, fit, test= "LRT")


calipers <- data.frame(
  group1 = c("late","late","late"),
  group2 = c("De Novo Metastatic","early","intermediate"),
  p = c(0.0017,0.012,0.092)
) |>
  filter(p < 0.05)



calipers$y.position <- c(0.55,0.65)
calipers$label <- sprintf("p = %.2g", calipers$p)


wgiicat<-ggplot(final, aes(x = categories_plot, y = wgii)) +
      geom_beeswarm(
    aes(color = site),size = 1.5,alpha=0.7,method="center")+
  geom_line(aes(y=predwgiicat,group=site,colour=site),linewidth = 1.2)+
    geom_boxplot(outlier.shape = NA, alpha = 0.3, width = 0.6) +

    stat_pvalue_manual(
        calipers,
        label = "label",
        tip.length = 0.01
    ) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"
                 ))+
    scale_x_discrete(
        labels = c(
            "De Novo Metastatic" = "DeNovo",
            "early" = "Early",
            "intermediate" = "Intermediate",
            "late" = "Late"
        )
    ) +
    labs(
        x = NULL,
        y = "WGII",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(0, 0.7)) +
    theme_minimal()


#continuous


fit <- lm(wgii ~ timetotrial + site, data = final)
summary(fit)
confint(fit)

cbind(
  coef(fit),
  confint(fit)
)

# Reduced model (no recurrence timing)
fit_reduced <- lm(wgii ~ site, data = final[!is.na(final$timetotrial),])

# Likelihood ratio test
anova(fit_reduced, fit, test= "LRT")


final$predwgii<-predict(fit,newdata = final)

# final$partial_angio <- resid(fit) + coef(fit)["timetotrial"] * final$timetotrial

p_label <- "p = 0.0096"

summary(fit)


final$timetotrialcut<-ifelse(final$timetotrial>120,120,final$timetotrial)

wgiicont <- ggplot(final, aes(x = timetotrialcut, y = wgii,colour=site)) +
    geom_point(alpha = 0.7) +geom_line(aes(x=timetotrial,y=predwgii),linewidth = 1.2) +
    annotate("text",
             x = 40,
             y = 0.55,
             label = p_label,
             hjust = 0,
             size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +  
    scale_color_manual(
        values = c(
            "KIDNEY" = "#4d9221",
            "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "WGII",
        colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(0,.7), xlim = c(0,120)) +
    theme_minimal()




#### TMB
#### TMB
#### TMB
#### TMB
#### TMB

#all analyses are done on log(TMB+1())
final$logtmb <- log10(final$tmb + 1)

fit <- lm(logtmb ~ categories_model + site, data = final)
final$predtmbcat<-predict(fit,newdata = final)
summary(fit)
cbind(
  coef(fit),
  confint(fit)
)


# Reduced model (no recurrence timing)
fit_reduced <- lm(logtmb ~ site, data = final)

# Likelihood ratio test
anova(fit_reduced, fit, test= "LRT")



calipers <- data.frame(
  group1 = c("late","late","late"),
  group2 = c("De Novo Metastatic","early","intermediate"),
  p = c(0.35,0.24,0.21)
) |>
  filter(p < 0.05)


calipers$label <- sprintf("p = %.2g", calipers$p)


tmbcat<-ggplot(final, aes(x = categories_plot, y = logtmb)) +
      geom_beeswarm(
    aes(color = site),size = 1.5,alpha=0.7,method="center")+
  geom_line(aes(y=predtmbcat,group=site,colour=site),linewidth = 1.2)+
    geom_boxplot(outlier.shape = NA, alpha = 0.3, width = 0.6) +


  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"
                 ))+
    scale_x_discrete(
        labels = c(
            "De Novo Metastatic" = "DeNovo",
            "early" = "Early",
            "intermediate" = "Intermediate",
            "late" = "Late"
        )
    ) +
    labs(
        x = NULL,
        y = "TMB",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(0, 2.6)) +
    theme_minimal()


#continuous


fit <- lm(logtmb ~ timetotrial + site, data = final)
summary(fit)
cbind(
  coef(fit),
  confint(fit)
)

# Reduced model (no recurrence timing)
fit_reduced <- lm(logtmb ~ site, data = final[!is.na(final$timetotrial),])

# Likelihood ratio test
anova(fit_reduced, fit, test= "LRT")


final$predtmb<-predict(fit,newdata = final)

# final$partial_angio <- resid(fit) + coef(fit)["timetotrial"] * final$timetotrial

p_label <- "p = 0.54"

summary(fit)

tmbcont <- ggplot(final, aes(x = timetotrialcut, y = logtmb,colour=site)) +
    geom_point(alpha = 0.7) +geom_line(aes(x=timetotrial,y=predtmb),linewidth = 1.2) +
    annotate("text",
             x = 40,
             y = 0.55,
             label = p_label,
             hjust = 0,
             size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +  
    scale_color_manual(
        values = c(
            "KIDNEY" = "#4d9221",
            "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "TMB",
        colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(0,2.5), xlim = c(0,120)) +
    theme_minimal()

combtmb<-plot_grid(
tmbcat,tmbcont,
  ncol = 2,rel_widths = c(1.3,1)
)

combonco<-plot_grid(
dntitle,etitle,ititle,ltitle,
  ncol = 4,rel_widths = c(1.3,1,1,1.5)
)

combwgii<-plot_grid(
wgiicat,wgiicont,
  ncol = 2,rel_widths = c(1.3,1)
)

mutplot<-plot_grid(
combonco,combwgii,combtmb,
  ncol = 1,rel_heights = c(1.8,1,1)
)

mutplot_buffered <- mutplot +
  theme(plot.margin = ggplot2::margin(t = 10, r = 0, b = 0, l = 10, unit = "mm"))
  
mutplot_buffered

```

## RNA analysis
```{r RNA}


natjavrna <- read_excel("/Users/smonda/Library/CloudStorage/OneDrive-MichiganMedicine/Research/M vs P/RDS121_2024_12_12T10_04_unzip/43018_2024_896_MOESM2_ESM(4).xlsx", 
                        sheet = "Supp. Table 3")
tpmjav<-data.frame(t(natjavrna))



colnames(tpmjav) <- natjavrna$hgnc_symbol

tpmjav<-tpmjav[-1,]

tpmjav_num <- data.frame(
  lapply(tpmjav, function(x) as.numeric(as.character(x))),
  row.names = rownames(tpmjav),
  check.names = FALSE
)
rm(tpmjav)
rm(natjavrna)

tpmjav_z <- as.data.frame(scale((tpmjav_num)))


g15<-c("IGF2BP3","PTK6","BOLA3","UQCRH","TFCP2L1","SLC7A8","SLC16A1","PLIN5", "SLC7A1","TMEM81","RPL22L1","TCEB2","IQCC","TAGLN2","TMEM136")

teff<-c("CD8A","IFNG","PRF1","EOMES","CD274")

teff %in% colnames(tpmjav_z)

cellcycle <- c("CDK2","CDK4","CDK6","BUB1","BUB1B","CCNE1",	"POLQ",	"AURKA",
         "MKI67","CCNB2")

cellcycle %in% colnames(tpmjav_z)

stroma<-c("FAP","FN1","COL5A1","COL5A2","POSTN","COL1A1","COL1A2","MMP2")
stroma %in% colnames(tpmjav_z)

#PECAM1 not in JAV RNA data
angio<-c("VEGFA","KDR","ESM1","PECAM1","ANGPTL4","CD34")

angio %in% colnames(tpmjav_z)

#CXCL8 not in JAV RNA data
myeloid<-c("CXCL1","CXCL2","CXCL3","CXCL8","IL6","PTGS2")

myeloid %in% colnames(tpmjav_z)

vn<-c("APOLD1","EDNRB","NOS3","PPAP2B")
#PPAP2B is synonymous with PLPP3
ir<-c("CEACAM1","CX3CL1","CCL5")
cgd<-c("EIF4EBP1","TUBB2A","LMNB1")
il6<-c("IL6")
#prolaris
# ccp <- c("FOXM1","ASPM","TK1","PRC1","CDC20",
#          "BUB1B",	"PBK",	"DTL",
#          "CDKN3","RRM2","ASF1B","CEP55",
#          "CDK1","DLGAP5","SKA1","RAD51",
#          "KIF11","BIRC5","RAD54L","CENPM",
#          "KIAA0101","KIF20A","PTTG1","CDCA8",
#          "NUSAP1","PLK1","CDCA3","ORC6",
#          "CENPF","TOP2A","MCM10")


#1.5 ╬╝g/mL
# 383.3 g/mol

#udager vince salami
emt<-c("LRRC15","GAS1","COL11A1","PTX3","COL7A1","COL8A2","LOXL2","SFRP4","PCOLCE","MXRA5","LOXL1","DKK1","CTHRC1","COL6A3","IL6","PRRX1","COL1A1","FBN2","SERPINE1","COMP","TFPI2","LUM")
emt %in% colnames(tpmjav_z)
# pam50prolif<-c("BIRC5", "CCNB1", "CDC20", "CDCA1", "CEP55", "KNTC2", "MKI67", "PTTG1", "RRM2", "TYMS", "UBE2C")

tpmjav_z$angio <- rowMedians(
  as.matrix(tpmjav_z[, colnames(tpmjav_z) %in% angio])
)

tpmjav_z$g15 <- rowSums(
  as.matrix(tpmjav_z[, colnames(tpmjav_z) %in% g15])
)


tpmjav_z$teff <- rowMedians(
  as.matrix(tpmjav_z[, colnames(tpmjav_z) %in% teff])
)

tpmjav_z$myeloid <- rowMedians(
  as.matrix(tpmjav_z[, colnames(tpmjav_z) %in% myeloid])
)


tpmjav_z$stroma <- rowMedians(
  as.matrix(tpmjav_z[, colnames(tpmjav_z) %in% stroma])
)


tpmjav_z$cellcycle <- rowMedians(
  as.matrix(tpmjav_z[, colnames(tpmjav_z) %in% cellcycle])
)


tpmjav_z$emt <- rowSums(
  as.matrix(tpmjav_z[, colnames(tpmjav_z) %in% emt])
)


final$angio<-tpmjav_z$angio[match(final$natid,rownames(tpmjav_z))]
final$cellcycle<-tpmjav_z$cellcycle[match(final$natid,rownames(tpmjav_z))]
final$teff<-tpmjav_z$teff[match(final$natid,rownames(tpmjav_z))]
final$myeloid<-tpmjav_z$myeloid[match(final$natid,rownames(tpmjav_z))]
final$stroma<-tpmjav_z$stroma[match(final$natid,rownames(tpmjav_z))]
final$emt<-tpmjav_z$emt[match(final$natid,rownames(tpmjav_z))]
final$g15<-tpmjav_z$g15[match(final$natid,rownames(tpmjav_z))]
final$igf2bp3<-tpmjav_z$IGF2BP3[match(final$natid,rownames(tpmjav_z))]


final$categories <- factor(final$categories)
final$categories <- relevel(final$categories, ref = "late")



#i would just focus on angio cell cycle tll myeloid teff, stroma is hard to interpret
#angion
#angio
#angio


fit <- lm(angio ~ categories_model + site, data = final)
cbind(
  coef(fit),
  confint(fit)
)

final$predangiocat<-predict(fit,newdata = final)
summary(fit)
# Reduced model (no recurrence timing)
fit_reduced <- lm(angio ~ site, data = final)
anova(fit_reduced, fit, test= "LRT")




calipers <- data.frame(
  group1 = c("late","late","late"),
  group2 = c("De Novo Metastatic","early","intermediate"),
  p = c(1.8e-06,3.3e-06,0.024)
) |>
  filter(p < 0.05)



calipers$y.position <- c(2.1,2.5,2.9)
calipers$label <- sprintf("p = %.2g", calipers$p)


angiocat<-ggplot(final, aes(x = categories_plot, y = angio)) +
      geom_beeswarm(
    aes(color = site),size = 1.5,alpha=0.7,method="center")+
  geom_line(aes(y=predangiocat,group=site,colour=site),linewidth = 1.2)+
    geom_boxplot(outlier.shape = NA, alpha = 0.3, width = 0.6) +

    stat_pvalue_manual(
        calipers,
        label = "label",
        tip.length = 0.01
    ) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"
                 ))+
    scale_x_discrete(
        labels = c(
            "De Novo Metastatic" = "DeNovo",
            "early" = "Early",
            "intermediate" = "Intermediate",
            "late" = "Late"
        )
    ) +
    labs(
        x = NULL,
        y = "Angiogenesis",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3, 3)) +
    theme_minimal()





fit <- lm(angio ~ timetotrial + site, data = final)
# fit <- lm(emt ~ timetotrial + site, data = final)

summary(fit)
cbind(
  coef(fit),
  confint(fit)
)

fit_reduced <- lm(angio ~ site, data = final[!is.na(final$timetotrial),])
anova(fit_reduced, fit, test= "LRT")


final$predangio<-predict(fit,newdata = final)

# final$partial_angio <- resid(fit) + coef(fit)["timetotrial"] * final$timetotrial

p_label <- "p = 0.00061"

summary(fit)

angiocont <- ggplot(final, aes(x = timetotrialcut, y = angio,colour=site)) +
    geom_point(alpha = 0.7) +geom_line(aes(x=timetotrial,y=predangio),linewidth = 1.2) +
    annotate("text",
             x = 40,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +  
    scale_color_manual(
        values = c(
            "KIDNEY" = "#4d9221",
            "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "Angiogenesis",
        colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
    theme_minimal()

###
###cell cycle

###
###cell cycle

###
###cell cycle

fit <- lm(cellcycle ~ categories_model + site, data = final)
final$predcellcyclecat<-predict(fit,newdata = final)
summary(fit)

cbind(
  coef(fit),
  confint(fit)
)

fit_reduced <- lm(cellcycle ~ site, data = final)
anova(fit_reduced, fit, test= "LRT")


calipers <- data.frame(
  group1 = c("late","late","late"),
  group2 = c("De Novo Metastatic","early","intermediate"),
  p = c(0.0019,0.0045,0.00015)
) |>
  filter(p < 0.05)



calipers$y.position <- c(2.1,2.5,2.9)
calipers$label <- sprintf("p = %.2g", calipers$p)


cellcyclecat<-ggplot(final, aes(x = categories_plot, y = cellcycle)) +
      geom_beeswarm(aes(color = site),size = 1.5,alpha=0.7,method="center") +
    geom_line(aes(y=predcellcyclecat,group=site,colour=site),linewidth = 1.2)+
  geom_boxplot(outlier.shape = NA, alpha = 0.3, width = 0.6) +
    stat_pvalue_manual(
        calipers,
        label = "label",
        tip.length = 0.01
    ) +
    scale_x_discrete(
        labels = c(
            "De Novo Metastatic" = "DeNovo",
            "early" = "Early",
            "intermediate" = "Intermediate",
            "late" = "Late"
        )
    ) +  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"
                 ))+
    labs(
        x = NULL,
        y = "Cell Cycle",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3, 3)) +
    theme_minimal()




fit <- lm(cellcycle ~ timetotrial + site, data = final)
final$predcellcycle<-predict(fit,newdata = final)
cbind(
  coef(fit),
  confint(fit)
)
fit_reduced <- lm(cellcycle ~ site, data = final[!is.na(final$timetotrial),])
anova(fit_reduced, fit, test= "LRT")

summary(fit)



p_label <- "p = 5.8e-06"


cellcyclecont <- ggplot(final, aes(x = timetotrialcut, y = cellcycle,colour=site)) +
    geom_point(alpha = 0.7) +
  geom_line(aes(x=timetotrial,y=predcellcycle),linewidth = 1.2)+
    annotate("text",
             x = 40,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"        ))+
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +
    labs(
        x = "Recurrence Time (months)",
        y = "Cell Cycle",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
    theme_minimal()

###

fit <- lm(cellcycle ~ categories_model + site, data = final)

cbind(
  coef(fit),
  confint(fit)
)

fit_reduced <- lm(cellcycle ~ site, data = final)
anova(fit_reduced, fit, test= "LRT")

fit <- lm(cellcycle ~ timetotrial + site, data = final)
fit_reduced <- lm(cellcycle ~ site, data = final[!is.na(final$timetotrial),])
anova(fit_reduced, fit, test= "LRT")

####

fit <- lm(myeloid ~ categories_model + site, data = final)
final$predmyeloidcat<-predict(fit,newdata = final)
summary(fit)

calipers <- data.frame(
  group1 = c("late","late","late"),
  group2 = c("De Novo Metastatic","early","intermediate"),
  p = c(0.57,0.06,0.15)
) |>
  filter(p < 0.05)



# calipers$y.position <- c(2.1,2.5,2.9)
# calipers$label <- sprintf("p = %.2g", calipers$p)


myeloidcat<-ggplot(final, aes(x = categories_plot, y = myeloid)) +
      geom_beeswarm(aes(color = site),size = 1.5,alpha=0.7,method="center") +
    geom_line(aes(y=predmyeloidcat,group=site,colour=site),linewidth = 1.2)+
  geom_boxplot(outlier.shape = NA, alpha = 0.3, width = 0.6) +
    scale_x_discrete(
        labels = c(
            "De Novo Metastatic" = "DeNovo",
            "early" = "Early",
            "intermediate" = "Intermediate",
            "late" = "Late"
        )
    ) +  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"
                 ))+
    labs(
        x = NULL,
        y = "Myeloid",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3, 3)) +
    theme_minimal()

myeloidcat


fit <- lm(myeloid ~ timetotrial + site, data = final)
final$predmyeloid<-predict(fit,newdata = final)
cbind(
  coef(fit),
  confint(fit)
)
fit_reduced <- lm(myeloid ~ site, data = final[!is.na(final$timetotrial),])
anova(fit_reduced, fit, test= "LRT")

summary(fit)



 p_label <- "p = 0.0008"


myeloidcont <- ggplot(final, aes(x = timetotrialcut, y = myeloid,colour=site)) +
    geom_point(alpha = 0.7) +
  geom_line(aes(x=timetotrial,y=predmyeloid),linewidth = 1.2)+
    annotate("text",
             x = 40,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"        ))+
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +
    labs(
        x = "Recurrence Time (months)",
        y = "Myeloid",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
    theme_minimal()


fit <- lm(myeloid ~ categories_model + site, data = final)
fit_reduced <- lm(myeloid ~ site, data = final)
anova(fit_reduced, fit, test= "LRT")

cbind(
  coef(fit),
  confint(fit)
)



fit <- lm(myeloid ~ timetotrial + site, data = final)
summary(fit)



cbind(
  coef(fit),
  confint(fit)
)

fit_reduced <- lm(myeloid ~ site, data = final[!is.na(final$timetotrial),])
anova(fit_reduced, fit, test= "LRT")

cbind(
  coef(fit),
  confint(fit)
)

####

fit <- lm(teff ~ categories_model + site, data = final)
final$predteffcat<-predict(fit,newdata = final)
summary(fit)

calipers <- data.frame(
  group1 = c("late","late","late"),
  group2 = c("De Novo Metastatic","early","intermediate"),
  p = c(0.55,0.24,0.36)
) |>
  filter(p < 0.05)



# calipers$y.position <- c(2.1,2.5,2.9)
# calipers$label <- sprintf("p = %.2g", calipers$p)
# 

teffcat<-ggplot(final, aes(x = categories_plot, y = teff)) +
      geom_beeswarm(aes(color = site),size = 1.5,alpha=0.7,method="center") +
    geom_line(aes(y=predteffcat,group=site,colour=site),linewidth = 1.2)+
  geom_boxplot(outlier.shape = NA, alpha = 0.3, width = 0.6) +
    scale_x_discrete(
        labels = c(
            "De Novo Metastatic" = "DeNovo",
            "early" = "Early",
            "intermediate" = "Intermediate",
            "late" = "Late"
        )
    ) +  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"
                 ))+
    labs(
        x = NULL,
        y = "T-effector",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3, 3)) +
    theme_minimal()

teffcat


fit <- lm(teff ~ timetotrial + site, data = final)
final$predteff<-predict(fit,newdata = final)
cbind(
  coef(fit),
  confint(fit)
)
fit_reduced <- lm(teff ~ site, data = final[!is.na(final$timetotrial),])
anova(fit_reduced, fit, test= "LRT")

summary(fit)



# p_label <- "p = 1.4e-06"


teffcont <- ggplot(final, aes(x = timetotrialcut, y = teff,colour=site)) +
    geom_point(alpha = 0.7) +
  geom_line(aes(x=timetotrial,y=predteff),linewidth = 1.2)+
    # annotate("text",
    #          x = 40,
    #          y = 2.2,
    #          # label = p_label,
    #          hjust = 0,
    #          size = 4) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"        ))+
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +
    labs(
        x = "Recurrence Time (months)",
        y = "T-effector",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
    theme_minimal()




fit <- lm(teff ~ categories_model + site, data = final)
cbind(
  coef(fit),
  confint(fit)
)
fit_reduced <- lm(teff ~ site, data = final)
anova(fit_reduced, fit, test= "LRT")

fit <- lm(teff ~ timetotrial + site, data = final)

cbind(
  coef(fit),
  confint(fit)
)
summary(fit)
fit_reduced <- lm(teff ~ site, data = final[!is.na(final$timetotrial),])
anova(fit_reduced, fit, test= "LRT")


####

fit <- lm(stroma ~ categories_model + site, data = final)
final$predstromacat<-predict(fit,newdata = final)
summary(fit)

calipers <- data.frame(
  group1 = c("late","late","late"),
  group2 = c("De Novo Metastatic","early","intermediate"),
  p = c(0.0028,0.00032,0.00003)
) |>
  filter(p < 0.05)



calipers$y.position <- c(2.1,2.5,2.9)
calipers$label <- sprintf("p = %.2g", calipers$p)


stromacat<-ggplot(final, aes(x = categories_plot, y = stroma)) +
      geom_beeswarm(aes(color = site),size = 1.5,alpha=0.7,method="center") +
    geom_line(aes(y=predstromacat,group=site,colour=site),linewidth = 1.2)+
  geom_boxplot(outlier.shape = NA, alpha = 0.3, width = 0.6) +
    stat_pvalue_manual(
        calipers,
        label = "label",
        tip.length = 0.01
    ) +
    scale_x_discrete(
        labels = c(
            "De Novo Metastatic" = "DeNovo",
            "early" = "Early",
            "intermediate" = "Intermediate",
            "late" = "Late"
        )
    ) +  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"
                 ))+
    labs(
        x = NULL,
        y = "Stroma",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3, 3)) +
    theme_minimal()




fit <- lm(stroma ~ timetotrial + site, data = final)
final$predstroma<-predict(fit,newdata = final)
cbind(
  coef(fit),
  confint(fit)
)
fit_reduced <- lm(stroma ~ site, data = final[!is.na(final$timetotrial),])
anova(fit_reduced, fit, test= "LRT")

summary(fit)



p_label <- "p = 1.4e-06"


stromacont <- ggplot(final, aes(x = timetotrialcut, y = stroma,colour=site)) +
    geom_point(alpha = 0.7) +
  geom_line(aes(x=timetotrial,y=predstroma),linewidth = 1.2)+
    annotate("text",
             x = 40,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"        ))+
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +
    labs(
        x = "Recurrence Time (months)",
        y = "Stroma",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
    theme_minimal()


fit <- lm(stroma ~ categories_model + site, data = final)

cbind(
  coef(fit),
  confint(fit)
)

fit_reduced <- lm(stroma ~ site, data = final)
anova(fit_reduced, fit, test= "LRT")

fit <- lm(stroma ~ timetotrial + site, data = final)
summary(fit)
cbind(
  coef(fit),
  confint(fit)
)

fit_reduced <- lm(stroma ~ site, data = final[!is.na(final$timetotrial),])
anova(fit_reduced, fit, test= "LRT")



###
###
# ### EMT
# 
# fit <- lm(emt ~ categories_model + site, data = final)
# summary(fit)
# 
# calipers <- data.frame(
#   group1 = c("late","late","late"),
#   group2 = c("De Novo Metastatic","early","intermediate"),
#   p = c(0.000746,8.26e-05,0.000117)
# ) |>
#   filter(p < 0.05)
# 
# 
# 
# calipers$y.position <- c(9.1,9.5,9.9)
# calipers$label <- sprintf("p = %.2g", calipers$p)
# 
# 
# emtcat<-ggplot(final, aes(x = categories_plot, y = emt)) +
#       geom_beeswarm(
#     aes(color = site),size = 1.5,alpha=0.7,method="center") +
#     geom_boxplot(outlier.shape = NA, alpha = 0.3, width = 0.6) +
# 
#     stat_pvalue_manual(
#         calipers,
#         label = "label",
#         tip.length = 0.01
#     ) +  
#   scale_color_manual(
#                  values = c(
#                      "KIDNEY" = "#4d9221",
#                      "METASTASIS" = "#c51b7d"
#                  ))+
#     scale_x_discrete(
#         labels = c(
#             "De Novo Metastatic" = "DeNovo",
#             "early" = "Early",
#             "intermediate" = "Intermediate",
#             "late" = "Late"
#         )
#     ) +
#     labs(
#         x = NULL,
#         y = "Angiogenesis",
#   colour = "Sequencing\nSite"
#     ) +
#     coord_cartesian(ylim = c(-10, 44)) +
#     theme_minimal()
# 
# 
# 
# 
# 
# fit <- lm(emt ~ timetotrial + site, data = final)
# 
# final$partial_emt <- resid(fit) + coef(fit)["timetotrial"] * final$timetotrial
# summary(fit)
# 
# p_label <- "p = 1.2e-6"
# 
# summary(fit)
# 
# emtcont <- ggplot(final, aes(x = timetotrialcut, y = emt,colour=site)) +
#     geom_point(alpha = 0.7) +
#     geom_smooth(method = "lm",   formula = y ~ x, se = TRUE, color = "black") +
#     annotate("text",
#              x = 40,
#              y = 2.2,
#              label = p_label,
#              hjust = 0,
#              size = 4) +
#     scale_x_continuous(
#         breaks = c(0,30,60,90,120),
#         labels = c("0","30","60","90",">120")
#     ) +  
#   scale_color_manual(
#                   values = c(
#                       "KIDNEY" = "#4d9221",
#                       "METASTASIS" = "#c51b7d"        ))+
#     labs(
#         x = "Recurrence Time (months)",
#         y = "Angiogenesis",
#   colour = "Sequencing\nSite"
#     ) +
#     coord_cartesian(ylim = c(-10,44), xlim = c(0,120)) +
#     theme_minimal()
# 
# 
# ###
# ###TLS
# ###
# ###TLS###
# ###TLS###
# ###TLS###
# ###TLS###
# ###TLS
# 
# 
# fit <- lm(tls ~ categories_model + site, data = final)
# summary(fit)
# 
# calipers <- data.frame(
#   group1 = c("late","late","late"),
#   group2 = c("De Novo Metastatic","early","intermediate"),
#   p = c(0.46,0.191,0.0355)
# ) |>
#   filter(p < 0.05)
# 
# 
# 
# calipers$y.position <- c(0.8)
# calipers$label <- sprintf("p = %.2g", calipers$p)
# 
# 
# tlscat<-ggplot(final, aes(x = categories_plot, y = tls)) +
#       geom_beeswarm(aes(color = site),size = 1.5,alpha=0.7,method="center") +  
#   geom_boxplot(outlier.shape = NA, alpha = 0.3, width = 0.6) +  
#   scale_color_manual(
#                  values = c(
#                      "KIDNEY" = "#4d9221",
#                      "METASTASIS" = "#c51b7d"
#                  ))+
#     stat_pvalue_manual(
#         calipers,
#         label = "label",
#         tip.length = 0.01
#     ) +
#     scale_x_discrete(
#         labels = c(
#             "De Novo Metastatic" = "DeNovo",
#             "early" = "Early",
#             "intermediate" = "Intermediate",
#             "late" = "Late"
#         )
#     ) +
#     labs(
#         x = NULL,
#         y = "TLS",
#   colour = "Sequencing\nSite"
#     ) +
#     coord_cartesian(ylim = c(0, 1)) +
#     theme_minimal()
# 
# 
# 
# 
# 
# fit <- lm(tls ~ timetotrial + site, data = final)
# summary(fit)
# final$partial_tls <- resid(fit) + coef(fit)["timetotrial"] * final$timetotrial
# summary(fit)
# p_label <- "p = 0.00014"
# 
# 
# 
# tlscont <- ggplot(final, aes(x = timetotrialcut, y = tls,colour=site)) +
#     geom_point(alpha = 0.7) +
#     geom_smooth(method = "lm",   formula = y ~ x, se = TRUE, color = "black") +
#     annotate("text",
#              x = 40,
#              y = 0.8,
#              label = p_label,
#              hjust = 0,
#              size = 4) +  scale_color_manual(
#                  values = c(
#                      "KIDNEY" = "#4d9221",
#                      "METASTASIS" = "#c51b7d"
#                  )
#              ) +
#     scale_x_continuous(
#         breaks = c(0,30,60,90,120),
#         labels = c("0","30","60","90",">120")
#     ) +
#     labs(
#         x = "Recurrence Time (months)",
#         y = "TLS",
#   colour = "Sequencing\nSite"
#     ) +
#     coord_cartesian(ylim = c(0,1), xlim = c(0,120)) +
#     theme_minimal()
# 
# 
# tlscont <- ggplot(final, aes(x = timetotrialcut, y = partial_tls)) +
#     geom_point(alpha = 0.7) +
#     geom_smooth(method = "lm", se = FALSE, color = "#00274C") +
#     annotate("text",
#              x = 40,
#              y = 0.8,
#              label = p_label,
#              hjust = 0,
#              size = 4) +
#     scale_x_continuous(
#         breaks = c(0,30,60,90,120),
#         labels = c("0","30","60","90",">120")
#     ) +
#     labs(
#         x = "Recurrence Time (months)",
#         y = "TLS"
#     ) +
#     coord_cartesian(ylim = c(0,1), xlim = c(0,120)) +
#     theme_minimal()
###
###



summary(lm(tls ~ categories + site, data = final))
summary(lm(teff ~ categories + site, data = final))
summary(lm(stroma ~ categories + site, data = final))
summary(lm(myeloid ~ categories + site, data = final))
summary(lm(cellcycle ~ categories + site, data = final))


# 
# final$igf2bp3<-tpmjav_z$IGF2BP3[match(final$natid,rownames(tpmjav_z))]
# final$hif2a<-tpmjav_z$EPAS1[match(final$natid,rownames(tpmjav_z))]
# final$hif1a<-tpmjav_z$HIF1A[match(final$natid,rownames(tpmjav_z))]
# final$havcr1<-tpmjav_z$HAVCR1[match(final$natid,rownames(tpmjav_z))]

# # final$pam50prolif<-tpmjav_z$pam50prolif[match(final$natid,rownames(tpmjav_z))]
# final$ki67<-tpmjav_z$MKI67[match(final$natid,rownames(tpmjav_z))]
# 
#only final with expression data
final$categories <- relevel(final$categories, ref = "late")

finalexp<-subset(final,final$natid %in% rownames(tpmjav_z))


design <- model.matrix(~ categories
                       +site
                       , data = finalexp)

ftpm<-tpmjav_num[match(finalexp$natid,rownames(tpmjav_num)),]

fit <- lmFit(t(ftpm), design)   # t() because limma expects genes x samples
fit <- eBayes(fit)

rank_dn <- sort(fit$t[, "categoriesDe Novo Metastatic"], decreasing = TRUE)
rank_early <- sort(fit$t[, "categoriesearly"], decreasing = TRUE)
rank_inter <- sort(fit$t[, "categoriesintermediate"], decreasing = TRUE)

hallmark <- msigdbr(species = "Homo sapiens", category = "H")
hallmark_list <- split(hallmark$gene_symbol, hallmark$gs_name)

fg_dn <- fgsea(
  pathways = hallmark_list,
  stats = rank_dn, nperm = 100000
)

fg_early <- fgsea(
  pathways = hallmark_list,
  stats = rank_early, nperm = 100000
)

fg_inter <- fgsea(
  pathways = hallmark_list,
  stats = rank_inter, nperm = 100000
)

fg_dn_df <- as.data.frame(fg_dn)
fg_early_df <- as.data.frame(fg_early)
fg_inter_df <- as.data.frame(fg_inter)


sig_dn   <- fg_dn_df$pathway[fg_dn_df$padj < 0.05]
sig_early <- fg_early_df$pathway[fg_early_df$padj < 0.05]
sig_inter <- fg_inter_df$pathway[fg_inter_df$padj < 0.05]

pathways_keep <- Reduce(intersect, list(sig_dn, sig_early, sig_inter))
pathways_keep_clean <- gsub("HALLMARK_", "", pathways_keep)

nes_mat <- data.frame(
  pathway = pathways_keep,
  DeNovo = fg_dn$NES[match(pathways_keep, fg_dn$pathway)],
  Early = fg_early$NES[match(pathways_keep, fg_early$pathway)],
  Intermediate = fg_inter$NES[match(pathways_keep, fg_inter$pathway)]
)


nes_mat <- nes_mat[order(nes_mat$Early+nes_mat$DeNovo+nes_mat$Intermediate, decreasing = FALSE), ]

nes_mat$Late <- 0

nes_mat$pathway_clean <- gsub("HALLMARK_", "", nes_mat$pathway)

nes_long <- melt(nes_mat, id.vars = c("pathway","pathway_clean"))

nes_long$pathway_clean <- factor(
  nes_long$pathway_clean,
  levels = nes_mat$pathway_clean
)

nes_long$pathway <- factor(nes_long$pathway, levels = nes_mat$pathway)


hallmarkheat <- ggplot(nes_long, aes(x = variable, y = pathway_clean, fill = value)) +
  geom_tile() +
  geom_text(
    data = subset(nes_long, variable!= "Late"),
    aes(label = "*"),
    size = 5,
    fontface = "bold",
    vjust = 0.75   # nudges asterisk down to visually center it
  ) +
    geom_text(
    data = subset(nes_long, variable== "Late"),
    aes(label = "REF"),
    size = 3,
    vjust = 0.75   # nudges asterisk down to visually center it
  ) +
  scale_fill_gradient2(
    low = "#2166ac", mid = "gray95", high = "#b2182b",
    midpoint = 0, limits = c(-2.1, 4.3), name = "NES"
  ) +
  theme_minimal() +
  scale_y_discrete(expand = expansion(mult = c(0.1, 0.0))) +
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0)) +
  labs(x = "", y = "",
       title = "Hallmark Pathway Enrichment Across Recurrence Groups-\nJAVELIN Renal 101 (n=599)")

hallmarkheat


####
########
####v
####
####
####

finalexp<-subset(final,final$natid %in% rownames(tpmjav_z))
finalexp <- subset(finalexp, categories != "De Novo Metastatic")
finalexp$timetotrial <- as.numeric(finalexp$timetotrial)

design <- model.matrix(~ timetotrial+site
                       
                       , data = finalexp)

ftpm<-tpmjav_num[match(finalexp$natid,rownames(tpmjav_z)),]

fit_time <- lmFit(t(ftpm), design)
fit_time <- eBayes(fit_time)


rank_time <- fit_time$t[, "timetotrial"]
names(rank_time) <- rownames(fit_time)

rank_time <- sort(rank_time, decreasing = TRUE)



fg_time <- fgsea(
  pathways = hallmark_list,
  stats = rank_time, nperm = 100000
)

fg_time_df <- as.data.frame(fg_time)

 sig_time   <- fg_time_df$pathway[fg_time_df$padj < 0.05]

fg_time_plot <- fg_time_df[match(pathways_keep, fg_time_df$pathway), ]

fg_time_plot$order<-match(fg_time_plot$pathway,nes_mat$pathway)

fg_time_plot$pathway_clean <- gsub("HALLMARK_", "", fg_time_plot$pathway)


fg_time_plot$nesflip<-(-fg_time_plot$NES)

hallmarkbar <- ggplot(fg_time_plot,
                      aes(x = nesflip, y = reorder(pathway_clean, order), fill = nesflip)) +
  geom_col() +
    geom_text(
    data = subset(fg_time_plot, padj < 0.05),
    aes(label = "*",
        hjust = ifelse(nesflip > 0, -0.3, 1.3)),
    size = 6,
    fontface = "bold"
  ) +
  scale_x_reverse() +
  scale_y_discrete(expand = expansion(mult = c(0.1, 0.0))) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "gray95",
    high = "#b2182b",
    midpoint = 0,
        limits = c(-2.1, 4.3),
    name = "NES"
  ) +
  theme_minimal()  +
    annotate(
        "text",
        x = 0,
        y = 0,   # adjust depending on spacing
        label = "ŌåÉ Enriched Earlier     Enriched Later ŌåÆ",
        vjust = 1,  size = 3
    )+
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0)
  )+
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50")+
  labs(
    x = "",
    y = "",
    title = "Hallmark Pathway Enrichment Across Recurrence Time-\nJAVELIN Renal 101 (recurrent only; n=360)"
  )+theme(legend.position = "none")


hallmarkbar

combined_heat <- plot_grid(
  hallmarkheat,
  hallmarkbar,
  ncol = 2,
  rel_widths = c(1.3, 1)
)

combined_heat

```


```{r S-TRAC validation}
clinstrac <- read_delim("~/Downloads/15194361/A6181109.clinical.unified4_meta.tsv", 
                   delim = "\t", escape_double = FALSE, 
                   trim_ws = TRUE)

clinmore <- read_delim("~/Downloads/15194361/A6181109_NC_figure_data_202207/Figure5-S2-S3_ElasticNet/A6181109/A6181109.20190807.clinicalv2_meta.tsv", 
                       delim = "\t", escape_double = FALSE, 
                       trim_ws = TRUE)
TPMr <- read_delim("~/Downloads/15194361/A6181109.TPM.qc.map.median.log.tsv", 
                   delim = "\t", escape_double = FALSE, 
                   trim_ws = TRUE)

rownames<-TPMr$HUGO
TPMr<-TPMr[,colnames(TPMr)!="HUGO"]
TPMr<-data.frame(TPMr)
rownames(TPMr)<-rownames
#log2tpm
TPM<-data.frame(t(TPMr))
TPMz<-data.frame(scale(TPM))
rm(TPMr)

clinmore$recur<-(clinmore$EVENTYP=="Disease Recurrence or occurrence of a Secondary Malignancy")
clinmore$recurtime<-ifelse(clinmore$recur,
                           ifelse(clinmore$EEVALUED<365,"early",
                                  ifelse(clinmore$EEVALUED<1095,"mod","lateish")),NA
                           )
clinmore$id <- sub("^A6181109\\s+", "", clinmore$row_id)

#subset to recurrences only
clinmores<-clinmore[clinmore$EVENTYP=="Disease Recurrence or occurrence of a Secondary Malignancy"&!is.na(clinmore$EVENTYP),]


TPMz$angio <- rowMedians(
  as.matrix(TPMz[, colnames(TPMz) %in% angio])
)


TPMz$teff <- rowMedians(
  as.matrix(TPMz[, colnames(TPMz) %in% teff])
)

TPMz$myeloid <- rowMedians(
  as.matrix(TPMz[, colnames(TPMz) %in% myeloid])
)


TPMz$stroma <- rowMedians(
  as.matrix(TPMz[, colnames(TPMz) %in% stroma])
)


TPMz$cellcycle <- rowMedians(
  as.matrix(TPMz[, colnames(TPMz) %in% cellcycle])
)


TPMz$emt <- rowSums(
  as.matrix(TPMz[, colnames(TPMz) %in% emt])
)


clinmores$angio<-TPMz$angio[match(clinmores$id,rownames(TPMz))]
clinmores$cellcycle<-TPMz$cellcycle[match(clinmores$id,rownames(TPMz))]
clinmores$teff<-TPMz$teff[match(clinmores$id,rownames(TPMz))]
clinmores$myeloid<-TPMz$myeloid[match(clinmores$id,rownames(TPMz))]
clinmores$stroma<-TPMz$stroma[match(clinmores$id,rownames(TPMz))]
clinmores$emt<-TPMz$emt[match(clinmores$id,rownames(TPMz))]
clinmores$time<-clinmores$EEVALUEM #months

##angio figure
fit <- lm(angio ~ time, data = clinmores)
summary(fit)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")


p_label <- "p = 0.035"

summary(fit)

angiostraccont <- ggplot(clinmores, aes(x = time, y = angio)) +
    geom_point(alpha = 0.7,colour="#4d9221") +
    geom_smooth(method = "lm",   formula = y ~ x, se = FALSE, color = "#4d9221") +
    annotate("text",
             x = 20,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60),
        labels = c("0","30","60")
    ) +  
  # scale_color_manual(
  #                 values = c(
  #                     "KIDNEY" = "#4d9221",
  #                     "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "Angiogenesis"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,75)) +
    theme_minimal()

##cellcycle figure
fit <- lm(cellcycle ~ time, data = clinmores)
summary(fit)

anova(fit, test= "LRT")

p_label <- "p = 0.063"

summary(fit)

##cellcycle
cellcyclestraccont <- ggplot(clinmores, aes(x = time, y = cellcycle)) +
    geom_point(alpha = 0.7,colour="#4d9221") +
    geom_smooth(method = "lm",   formula = y ~ x, se = FALSE, color = "#4d9221") +
    annotate("text",
             x = 20,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60),
        labels = c("0","30","60")
    ) +  
  # scale_color_manual(
  #                 values = c(
  #                     "KIDNEY" = "#4d9221",
  #                     "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "Cell Cycle"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,75)) +
    theme_minimal()


##cellcycle figure
fit <- lm(stroma ~ time, data = clinmores)
summary(fit)

anova(fit, test= "LRT")

p_label <- "p = 0.023"

summary(fit)

##stroma
stromastraccont <- ggplot(clinmores, aes(x = time, y = stroma)) +
    geom_point(alpha = 0.7,colour="#4d9221") +
    geom_smooth(method = "lm",   formula = y ~ x, se = FALSE, color = "#4d9221") +
    annotate("text",
             x = 20,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60),
        labels = c("0","30","60")
    ) +  
  # scale_color_manual(
  #                 values = c(
  #                     "KIDNEY" = "#4d9221",
  #                     "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "Stroma"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,75)) +
    theme_minimal()


##myeloid figure
fit <- lm(myeloid ~ time, data = clinmores)
summary(fit)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")


# p_label <- "p = 0.55"

summary(fit)

myeloidstraccont <- ggplot(clinmores, aes(x = time, y = myeloid)) +
    geom_point(alpha = 0.7,colour="#4d9221") +
    geom_smooth(method = "lm",   formula = y ~ x, se = FALSE, color = "#4d9221") +
    # annotate("text",
    #          x = 20,
    #          y = 2.2,
    #          label = p_label,
    #          hjust = 0,
    #          size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60),
        labels = c("0","30","60")
    ) +  
  # scale_color_manual(
  #                 values = c(
  #                     "KIDNEY" = "#4d9221",
  #                     "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "Myeloid"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,75)) +
    theme_minimal()

##teff figure
fit <- lm(teff ~ time, data = clinmores)
summary(fit)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")


# p_label <- "p = 0.55"

summary(fit)

teffstraccont <- ggplot(clinmores, aes(x = time, y = teff)) +
    geom_point(alpha = 0.7,colour="#4d9221") +
    geom_smooth(method = "lm",   formula = y ~ x, se = FALSE, color = "#4d9221") +
    # annotate("text",
    #          x = 20,
    #          y = 2.2,
    #          label = p_label,
    #          hjust = 0,
    #          size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60),
        labels = c("0","30","60")
    ) +  
  # scale_color_manual(
  #                 values = c(
  #                     "KIDNEY" = "#4d9221",
  #                     "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "Myeloid"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,75)) +
    theme_minimal()



#hallmarkpathway

finalexp<-subset(clinmores,clinmores$id %in% rownames(TPM))
finalexp$timetotrial <- as.numeric(finalexp$EEVALUEM)

design <- model.matrix(~ timetotrial
                       
                       , data = finalexp)

ftpm<-TPM[match(finalexp$id,rownames(TPM)),]


fit_time <- lmFit(t(ftpm), design)
fit_time <- eBayes(fit_time)


rank_time <- fit_time$t[, "timetotrial"]
names(rank_time) <- rownames(fit_time)

rank_time <- sort(rank_time, decreasing = TRUE)


fg_time <- fgsea(
  pathways = hallmark_list,
  stats = rank_time, nperm = 100000
)

fg_time_df <- as.data.frame(fg_time)

sig_time   <- fg_time_df$pathway[fg_time_df$padj < 0.1]

fg_time_plot <- fg_time_df[match(pathways_keep, fg_time_df$pathway), ]

fg_time_plot$order<-match(fg_time_plot$pathway,nes_mat$pathway)

fg_time_plot$pathway_clean <- gsub("HALLMARK_", "", fg_time_plot$pathway)


fg_time_plot$nesflip<-(-fg_time_plot$NES)

hallmarkbarstrac <- ggplot(fg_time_plot,
                      aes(x = nesflip, y = reorder(pathway_clean, order), fill = nesflip)) +
  geom_col() +
  
  geom_text(
    data = subset(fg_time_plot, padj < 0.05),
    aes(label = "*",
        hjust = ifelse(nesflip > 0, -0.3, 1.3)),
    size = 6,
    fontface = "bold"
  ) +
  scale_x_reverse() +
  scale_y_discrete(expand = expansion(mult = c(0.1, 0.0))) +
  scale_fill_gradient2(
    low = "#2166ac", mid = "gray95", high = "#b2182b",
    midpoint = 0, limits = c(-2.1, 4.3), name = "NES"
  ) +
  theme_minimal() +
  annotate("text", x = 0, y = 0,
           label = "ŌåÉ Enriched Earlier     Enriched Later ŌåÆ",
           vjust = 1, size = 3) +
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0),
        legend.position = "none") +
  labs(x = "", y = "",
       title = "Hallmark Pathway Enrichment Across Recurrence Time-\nS-TRAC Recurrences (n=42)")


hallmarkbarstrac






```

```{r IMmotion 010 validation}

imm010clin <- read_excel("~/Library/CloudStorage/OneDrive-MichiganMedicine/Research/IMotion010/WO39210_ANON_IMmotion010.baseline.patient.data.778.20250513.xlsx")

#remove M1NED patients
imm010clinnephrec<-imm010clin[imm010clin$SCORE!="M1 NED",]

imm010fmi <- read_excel("~/Library/CloudStorage/OneDrive-MichiganMedicine/Research/IMotion010/WO39210_ANON_IMmotion010.baseline.recurrence.FMI.data.160.20250304.xlsx")


imm010clinrecur <- read_excel("~/Library/CloudStorage/OneDrive-MichiganMedicine/Research/IMotion010/WO39210_ANON_IMmotion010.baseline.recurrence.patient.data.160.20250513.xlsx")
#eliminates redundant
imm010clinrecur<-imm010clinrecur[imm010clinrecur$VISIT=="progression",]

imm010clinrecur$score<-imm010clin$SCORE[match(imm010clinrecur$HASHED_PATID,imm010clin$HASHED_PATID)]

imm010rnarecur <- read_excel("~/Library/CloudStorage/OneDrive-MichiganMedicine/Research/IMotion010/WO39210_ANON_IMmotion010.baseline.recurrence.rnaseq.160.20250304.xlsx")

imm010clinrecur<-imm010clinrecur[imm010clinrecur$VISIT=="progression",]

imm010clinrecur$score<-imm010clin$SCORE[match(imm010clinrecur$HASHED_PATID,imm010clin$HASHED_PATID)]


imm010rna <- read_excel("~/Library/CloudStorage/OneDrive-MichiganMedicine/Research/IMotion010/WO39210_ANON_IMmotion010.baseline.rnaseq.754.20250304.xlsx")

imm010rnarecur<-imm010rnarecur[,!colnames(imm010rnarecur) %in% colnames(imm010rna)]

#again could consider removing this
# CAN ONLY DO ONCE
 imm010rna<-cbind(imm010rna,imm010rnarecur)

#eliminating some unnamed genes (2500, could include for more granulatity)
imm010rna<-imm010rna[,colnames(imm010rna)!="Unnamed: 0"]
imm010rna<-imm010rna[!is.na(imm010rna$GENE_SYMBOL),]


genes<-imm010rna$GENE_SYMBOL


imm010trna<-as.data.frame(t(imm010rna))

colnames(imm010trna)<-imm010rna$GENE_SYMBOL
imm010trna<-imm010trna[rownames(imm010trna)!="GENE_SYMBOL",]
imm010trna[] <- lapply(imm010trna, as.numeric)

#removes non-nephrectomy patients
# eliminate M1 NED patients you are focusing on nephrectomy recurrence
#can scale before or after does not really matter, but for the sake of approach consistency did it after


# imm010trna<-imm010trna[rownames(imm010trna) %in% imm010clinnephrec$HASHED_RNA_SAMPLE_ID,]


imm010rnaz <- as.data.frame(scale(imm010trna))




imm010rnaz$angio <- rowMedians(
  as.matrix(imm010rnaz[, colnames(imm010rnaz) %in% angio])
)


imm010rnaz$teff <- rowMedians(
  as.matrix(imm010rnaz[, colnames(imm010rnaz) %in% teff])
)

imm010rnaz$myeloid <- rowMedians(
  as.matrix(imm010rnaz[, colnames(imm010rnaz) %in% myeloid])
)


imm010rnaz$g15 <- rowSums(
  as.matrix(imm010rnaz[, colnames(imm010rnaz) %in% g15])
)

imm010rnaz$stroma <- rowMedians(
  as.matrix(imm010rnaz[, colnames(imm010rnaz) %in% stroma])
)

imm010rnaz$cellcycle <- rowMedians(
  as.matrix(imm010rnaz[, colnames(imm010rnaz) %in% cellcycle])
)


imm010rnaz$emt <- rowSums(
  as.matrix(imm010rnaz[, colnames(imm010rnaz) %in% emt])
)

imm010rnaz$emtmed <- rowMedians(
  as.matrix(imm010rnaz[, colnames(imm010rnaz) %in% emt])
)

imm010rnaz$g15 <- rowSums(
  as.matrix(imm010rnaz[, colnames(imm010rnaz) %in% g15])
)

# imm010rnaz$hallmarkemt <- rowSums(
#   as.matrix(imm010rnaz[, colnames(imm010rnaz) %in% hallmarkemt])
# )

# 
# # Full table from the limma fit
# res <- topTable(fit_time, coef = "timetotrial",
#                 number = Inf, sort.by = "none")
# res$gene <- rownames(res)
# 
# 
# design <- model.matrix(~ timetotrial+site
# 
#                        , data = finalexp)
# 
# ftpm<-tpmjav_num[match(finalexp$natid,rownames(tpmjav_z)),]
# fit_time <- lmFit(t(ftpm), design)
# fit_time <- eBayes(fit_time)
# rank_time <- fit_time$t[, "timetotrial"]
# names(rank_time) <- rownames(fit_time)
# 
# # Full table from the limma fit
# res <- topTable(fit_time, coef = "timetotrial",
#                 number = Inf, sort.by = "none")
# res$gene <- rownames(res)
# 
# 
# # Top 25 genes upregulated in EARLY recurrence (most negative t)
# early_genes <- res[order(res$t), ][1:100, ]
# 
# # Top 25 genes upregulated in LATE recurrence (most positive t)
# late_genes <- res[order(-res$t), ][1:100, ]
# 
# early_genes[, c("gene","logFC","t","P.Value","adj.P.Val")]
# late_genes[,  c("gene","logFC","t","P.Value","adj.P.Val")]
# 
# 
# # imm010rnaz$tengenes <- rowSums(
# #   as.matrix(imm010rnaz[, colnames(imm010rnaz) %in% tengenes])
# # )
# late_mat <- as.matrix(imm010rnaz[, colnames(imm010rnaz) %in% rownames(late_genes)])
# late_score <- rowMeans(late_mat, na.rm = TRUE)
# 
# imm010rnaz$tengenes <-
#   rowMeans(
#   as.matrix(imm010rnaz[, colnames(imm010rnaz) %in% rownames(early_genes)])
# )-
# late_score
# # 
# # imm010clin$tengenes<-imm010rnaz$tengenes[match(imm010clin$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
# # 
# imm010clins<-imm010clin[imm010clin$SCORE!="M1 NED",]
# imm010clins$high<-imm010clins$emt>median(as.numeric(imm010clins$emt), na.rm = TRUE)
# imm010clins$highs<-imm010clins$emtmed>median(as.numeric(imm010clins$emtmed), na.rm = TRUE)
# 
# 
# 
# 
# # # imm010clins<-imm010clin
# 
# 
 # imm010clins<-subset(imm010clin,imm010clin$KIM1_C1D1>200)

# ggsurvplot(survfit(Surv(DFS_MONTHS,DFS_CENSOR)~
#                        TREATMENT_ARM
#                    ,
#                    data=imm010clins)
#            ,risk.table = TRUE,cumevents  =TRUE)


imm010clin$angio<-imm010rnaz$angio[match(imm010clin$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clin$cellcycle<-imm010rnaz$cellcycle[match(imm010clin$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clin$teff<-imm010rnaz$teff[match(imm010clin$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clin$myeloid<-imm010rnaz$myeloid[match(imm010clin$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clin$stroma<-imm010rnaz$stroma[match(imm010clin$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clin$emt<-imm010rnaz$emt[match(imm010clin$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clin$tengenes<-imm010rnaz$tengenes[match(imm010clin$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clin$g15<-imm010rnaz$g15[match(imm010clin$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
# imm010clin$hallmarkemt<-imm010rnaz$hallmarkemt[match(imm010clin$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clin$emtmed<-imm010rnaz$emtmed[match(imm010clin$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]

imm010clinrecur$angio<-imm010rnaz$angio[match(imm010clinrecur$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clinrecur$cellcycle<-imm010rnaz$cellcycle[match(imm010clinrecur$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clinrecur$teff<-imm010rnaz$teff[match(imm010clinrecur$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clinrecur$myeloid<-imm010rnaz$myeloid[match(imm010clinrecur$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clinrecur$stroma<-imm010rnaz$stroma[match(imm010clinrecur$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clinrecur$emt<-imm010rnaz$emt[match(imm010clinrecur$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
# imm010clinrecur$tengenes<-imm010rnaz$tengenes[match(imm010clinrecur$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clinrecur$g15<-imm010rnaz$g15[match(imm010clinrecur$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
# imm010clin$hallmarkemt<-imm010rnaz$hallmarkemt[match(imm010clin$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]
imm010clinrecur$emtmed<-imm010rnaz$emtmed[match(imm010clinrecur$HASHED_RNA_SAMPLE_ID,rownames(imm010rnaz))]

imm010clinrecur$time<-imm010clinrecur$DFS_MONTHS

imm010clin$angiorecur<-imm010clinrecur$angio[match(imm010clin$HASHED_PATID,imm010clinrecur$HASHED_PATID)]

imm010clin$nmfrecur<-imm010clinrecur$NMF_SUBTYPE[match(imm010clin$HASHED_PATID,imm010clinrecur$HASHED_PATID)]

imm010clin$time<-imm010clin$DFS_MONTHS





# ggplot(imm010clinrecurnephec, aes(x = time, y = emt,colour= TREATMENT_ARM)) +
#     geom_point(alpha = 0.7) 

# imm010clinrecurnephrecprim<-imm010clin[match(imm010clinrecurnephec$HASHED_PATID,imm010clin$HASHED_PATID),]

#non M1NED patients that recur on placebo

imm010clin$categories_plot<-ifelse(imm010clin$time<12,"early","intermediate")

imm010placeborec<-imm010clin[imm010clin$SCORE!="M1 NED" &imm010clin$TREATMENT_ARM=="PLACEBO" & imm010clin$DFS_CENSOR == TRUE,]

imm010atezorec<-imm010clin[imm010clin$SCORE!="M1 NED" &imm010clin$TREATMENT_ARM!="PLACEBO" & imm010clin$DFS_CENSOR == TRUE,]

imm010rec<-imm010clin[imm010clin$SCORE!="M1 NED" & imm010clin$DFS_CENSOR == TRUE,]

#this is recurrence sequencing
# imm010clinrecurnephrec<-imm010clinrecur[imm010clinrecur$score!="M1 NED",]
# 
# imm010clinrecurnephrec$angioprim<-imm010clin$angio[match(imm010clinrecurnephrec$HASHED_PATID,imm010clin$HASHED_PATID)]
# imm010clinrecurnephrec$NMF_SUBTYPE<-imm010clin$NMF_SUBTYPE[match(imm010clinrecurnephrec$HASHED_PATID,imm010clin$HASHED_PATID)]


##angio figure
fit <- lm(angio ~ time, data = imm010placeborec)
summary(fit)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")


p_label <- "p = 0.0014"

summary(fit)

angioimm010cont <- ggplot(imm010placeborec, aes(x = time, y = angio)) +
    geom_point(alpha = 0.7,colour="#4d9221") +
    geom_smooth(method = "lm",   formula = y ~ x, se = FALSE, color = "#4d9221") +
    annotate("text",
             x = 20,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60),
        labels = c("0","30","60")
    ) +  
  # scale_color_manual(
  #                 values = c(
  #                     "KIDNEY" = "#4d9221",
  #                     "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "Angiogenesis"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,60)) +
    theme_minimal()

##cellcycle figure
fit <- lm(cellcycle ~ time, data = imm010placeborec)
summary(fit)

cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")

p_label <- "p = 0.0045"

summary(fit)

##cellcycle
cellcycleimm010cont <- ggplot(imm010placeborec, aes(x = time, y = cellcycle)) +
    geom_point(alpha = 0.7,colour="#4d9221") +
    geom_smooth(method = "lm",   formula = y ~ x, se = FALSE, color = "#4d9221") +
    annotate("text",
             x = 20,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60),
        labels = c("0","30","60")
    ) +  
  # scale_color_manual(
  #                 values = c(
  #                     "KIDNEY" = "#4d9221",
  #                     "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "Cell Cycle"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,60)) +
    theme_minimal()


##stroma figure
fit <- lm(stroma ~ time, data = imm010placeborec)


summary(fit)

cbind(
  coef(fit),
  confint(fit)
)


anova(fit, test= "LRT")

p_label <- "p = 0.0049"

summary(fit)

##stroma
stromasimm010cont <- ggplot(imm010placeborec, aes(x = time, y = stroma)) +
    geom_point(alpha = 0.7,colour="#4d9221") +
    geom_smooth(method = "lm",   formula = y ~ x, se = FALSE, color = "#4d9221") +
    annotate("text",
             x = 20,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60),
        labels = c("0","30","60")
    ) +  
  # scale_color_manual(
  #                 values = c(
  #                     "KIDNEY" = "#4d9221",
  #                     "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "Stroma"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,60)) +
    theme_minimal()


##myeloid figure
fit <- lm(myeloid ~ time, data = imm010placeborec)
summary(fit)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")


 p_label <- "p = 0.037"

summary(fit)

myeloidimm010cont <- ggplot(imm010placeborec, aes(x = time, y = myeloid)) +
    geom_point(alpha = 0.7,colour="#4d9221") +
    geom_smooth(method = "lm",   formula = y ~ x, se = FALSE, color = "#4d9221") +
      annotate("text",
             x = 20,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +
    # annotate("text",
    #          x = 20,
    #          y = 2.2,
    #          label = p_label,
    #          hjust = 0,
    #          size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60),
        labels = c("0","30","60")
    ) +  
  # scale_color_manual(
  #                 values = c(
  #                     "KIDNEY" = "#4d9221",
  #                     "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "Myeloid"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,60)) +
    theme_minimal()

##teff figure
fit <- lm(teff ~ time, data = imm010placeborec)
summary(fit)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")


# p_label <- "p = 0.55"

summary(fit)

teffimm010cont <- ggplot(imm010placeborec, aes(x = time, y = teff)) +
    geom_point(alpha = 0.7,colour="#4d9221") +
    geom_smooth(method = "lm",   formula = y ~ x, se = FALSE, color = "#4d9221") +
    # annotate("text",
    #          x = 20,
    #          y = 2.2,
    #          label = p_label,
    #          hjust = 0,
    #          size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60),
        labels = c("0","30","60")
    ) +  
  # scale_color_manual(
  #                 values = c(
  #                     "KIDNEY" = "#4d9221",
  #                     "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "T-Effector"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,60)) +
    theme_minimal()



#hallmarkpathway

finalexp<-subset(imm010placeborec,imm010placeborec$HASHED_RNA_SAMPLE_ID %in% rownames(imm010trna))

# finalexp<-subset(imm010rec,imm010rec$HASHED_RNA_SAMPLE_ID %in% rownames(imm010trna))


# finalexp<-subset(imm010atezorec,imm010atezorec$HASHED_RNA_SAMPLE_ID %in% rownames(imm010trna))

# finalexp<-subset(imm010rec,imm010rec$HASHED_RNA_SAMPLE_ID %in% rownames(imm010trna))


finalexp$timetotrial <- as.numeric(finalexp$DFS_MONTHS)

# design <- model.matrix(~ TREATMENT_ARM
# 
#                        , data = finalexp)

design <- model.matrix(~ timetotrial
                       
                       , data = finalexp)

ftpm<-imm010trna[match(finalexp$HASHED_RNA_SAMPLE_ID,rownames(imm010trna)),]


fit_time <- lmFit(t(ftpm), design)
fit_time <- eBayes(fit_time)


rank_time <- fit_time$t[, "timetotrial"]


# rank_time <- fit_time$t[, "TREATMENT_ARMPLACEBO"]

names(rank_time) <- rownames(fit_time)

rank_time <- sort(rank_time, decreasing = TRUE)


rank_time <- rank_time[!is.na(rank_time)]
rank_time <- rank_time[order(-abs(rank_time))]
rank_time <- rank_time[!duplicated(names(rank_time))]
rank_time <- sort(rank_time, decreasing = TRUE)

fg_time <- fgsea(
  pathways = hallmark_list,
  stats = rank_time, nperm = 100000
)



fg_time_df <- as.data.frame(fg_time)

sig_time   <- fg_time_df$pathway[fg_time_df$padj < 0.1]

fg_time_plot <- fg_time_df[match(pathways_keep, fg_time_df$pathway), ]

fg_time_plot$order<-match(fg_time_plot$pathway,nes_mat$pathway)

fg_time_plot$pathway_clean <- gsub("HALLMARK_", "", fg_time_plot$pathway)


fg_time_plot$nesflip<-(-fg_time_plot$NES)

hallmarkbarimm010 <- ggplot(fg_time_plot,
                      aes(x = nesflip, y = reorder(pathway_clean, order), fill = nesflip)) +
  geom_col() +  
  
  scale_x_reverse() +
  scale_y_discrete(expand = expansion(mult = c(0.1, 0.0))) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "gray95",
    high = "#b2182b",
    midpoint = 0,
      limits = c(-2.1, 4.3),
    name = "NES"
  )  +
  theme_minimal()  +
    annotate(
        "text",
        x = 0,
        y = 0,   # adjust depending on spacing
        label = "ŌåÉ Enriched Earlier     Enriched Later ŌåÆ",
        vjust = 1,  size = 3
    ) +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0)
  )+
  labs(
    x = "",
    y = "",
    title = "Hallmark Pathway Enrichment Across Recurrence Time-\nIMmotion 010 Recurrences (placebo only; n=88)"
  )+theme(legend.position = "none")


hallmarkbarimm010
hallmarkbarimm010 <- ggplot(fg_time_plot,
                      aes(x = nesflip, y = reorder(pathway_clean, order), fill = nesflip)) +
  geom_col() +
  geom_text(
    data = subset(fg_time_plot, padj < 0.05),
    aes(label = "*",
        hjust = ifelse(nesflip > 0, -0.3, 1.3)),
    size = 6,
    fontface = "bold"
  ) +
  scale_x_reverse() +
  scale_y_discrete(expand = expansion(mult = c(0.1, 0.0))) +
  scale_fill_gradient2(
    low = "#2166ac", mid = "gray95", high = "#b2182b",
    midpoint = 0, limits = c(-2.1, 4.3), name = "NES"
  ) +
  theme_minimal() +
  annotate("text", x = 0, y = 0,
           label = "ŌåÉ Enriched Earlier     Enriched Later ŌåÆ",
           vjust = 1, size = 3) +
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0),
        legend.position = "none") +
  labs(x = "", y = "",
       title = "Hallmark Pathway Enrichment Across Recurrence Time-\nIMmotion 010 Recurrences (placebo only; n=88)")



combined_all<-plot_grid(
    hallmarkheat,
    hallmarkbar,angiocat,angiocont,cellcyclecat,cellcyclecont,
    ncol = 2,
    rel_widths = c(1.3, 1),rel_heights = c(2,1,1)
)

combined_allstrac<-plot_grid(
    hallmarkbarstrac,angiostraccont,cellcyclestraccont,
    ncol =1,rel_heights = c(2,1,1)
)

combinedjavstract<-plot_grid(
    hallmarkheat,
    hallmarkbar,hallmarkbarstrac,
    angiocat,angiocont,angiostraccont,cellcyclecat,cellcyclecont,cellcyclestraccont,
    ncol = 3,
    rel_widths = c(1, 1,1),rel_heights = c(1,0.6,0.6)
)

# combinedjavstracimm010<-plot_grid(
#     hallmarkheat,
#     hallmarkbar,hallmarkbarstrac,hallmarkbarimm010,
#     angiocat,angiocont,angiostraccont,angioimm010cont,
#     cellcyclecat,cellcyclecont,cellcyclestraccont,cellcycleimm010cont,
#     stromacat,stromacont,stromastraccont,stromasimm010cont,
#     ncol = 4,
#     rel_widths = c(1, 1,0.8,0.75),rel_heights = c(1,0.6,0.6)
# )

combinedjavstracimm010<-plot_grid(
    hallmarkheat,
    hallmarkbar,hallmarkbarstrac,hallmarkbarimm010,
    angiocat,angiocont,angiostraccont,angioimm010cont,
    cellcyclecat,cellcyclecont,cellcyclestraccont,cellcycleimm010cont,
    ncol = 4,
    rel_widths = c(1, 1,1,1),rel_heights = c(1,0.5,0.5)
)
combinedjavstracimm010

combinedjavstracimm010_buffered <- combinedjavstracimm010 +
  theme(plot.margin = ggplot2::margin(t = 10, r = 0, b = 0, l = 10, unit = "mm"))


combinedjavstract
combined_all

combinedstroma<-plot_grid(
stromacat,stromacont,stromastraccont,stromasimm010cont,
    ncol = 4,
    rel_widths = c(1, 1,1,1)
)

combinedteff<-plot_grid(
teffcat,teffcont,teffstraccont,teffimm010cont,
    ncol = 4,
    rel_widths = c(1, 1,1,1)
)


combinedmyeloid<-plot_grid(
myeloidcat,myeloidcont,myeloidstraccont,myeloidimm010cont,
    ncol = 4,
    rel_widths = c(1, 1,1,1)
)


```
#stratified analyses- metastasis sequencing only, primary tumor sequencing only
#and on co-author suggestion exclusion of sarcomatoid, and stratified to no sarcomatoid, and high grade (1/2) and low grade (3/4)

```{r sensitivity analysis to ensure significance is retained for only primary and only met site sequencing, applied on continuous analyses (early vs late, DEG)}

#metastasis only
finalexp<-subset(final,final$natid %in% rownames(tpmjav_z))
finalexp <- subset(finalexp, categories != "De Novo Metastatic")
finalexp<-subset(finalexp, site == "METASTASIS")
finalexp$timetotrial <- as.numeric(finalexp$timetotrial)


fit <- lm(wgii ~ timetotrial, data = finalexp)
finalexp$predwgii<-predict(fit,newdata = finalexp)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")

summary(fit)



p_label <- "p = 0.061"


wgiimetonly <- ggplot(finalexp, aes(x = timetotrialcut, y = wgii,colour=site)) +
    geom_point(alpha = 0.7) +
  geom_line(aes(x=timetotrial,y=predwgii),linewidth = 1.2)+
    annotate("text",
             x = 40,
             y = 0.5,
             label = p_label,
             hjust = 0,
             size = 4) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"        ))+
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +
    labs(
        x = "Recurrence Time (months)",
        y = "WGII",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(0,0.7), xlim = c(0,120)) +
    theme_minimal()


fit <- lm(cellcycle ~ timetotrial, data = finalexp)
finalexp$predcellcycle<-predict(fit,newdata = finalexp)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")

summary(fit)



p_label <- "p = 5e-5"


cellcyclemetonly <- ggplot(finalexp, aes(x = timetotrialcut, y = cellcycle,colour=site)) +
    geom_point(alpha = 0.7) +
  geom_line(aes(x=timetotrial,y=predcellcycle),linewidth = 1.2)+
    annotate("text",
             x = 40,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"        ))+
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +
    labs(
        x = "Recurrence Time (months)",
        y = "Cell Cycle",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
    theme_minimal()



fit <- lm(angio ~ timetotrial, data = finalexp)
finalexp$predangio<-predict(fit,newdata = finalexp)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")

summary(fit)



p_label <- "p = 0.16"


angiometonly <- ggplot(finalexp, aes(x = timetotrialcut, y = angio,colour=site)) +
    geom_point(alpha = 0.7) +
  geom_line(aes(x=timetotrial,y=predangio),linewidth = 1.2)+
    annotate("text",
             x = 40,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"        ))+
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +
    labs(
        x = "Recurrence Time (months)",
        y = "Angio",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
    theme_minimal()


fit <- lm(stroma ~ timetotrial, data = finalexp)
finalexp$predstroma<-predict(fit,newdata = finalexp)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")

summary(fit)



p_label <- "p = 0.00013"


stromametonly <- ggplot(finalexp, aes(x = timetotrialcut, y = stroma,colour=site)) +
    geom_point(alpha = 0.7) +
  geom_line(aes(x=timetotrial,y=predstroma),linewidth = 1.2)+
    annotate("text",
             x = 40,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"        ))+
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +
    labs(
        x = "Recurrence Time (months)",
        y = "Stroma",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
    theme_minimal()


#hallmark

design <- model.matrix(~ timetotrial
                       
                       , data = finalexp)

ftpm<-tpmjav_num[match(finalexp$natid,rownames(tpmjav_z)),]


fit_time <- lmFit(t(ftpm), design)
fit_time <- eBayes(fit_time)


rank_time <- fit_time$t[, "timetotrial"]
names(rank_time) <- rownames(fit_time)

rank_time <- sort(rank_time, decreasing = TRUE)


fg_time <- fgsea(
  pathways = hallmark_list,
  stats = rank_time, nperm = 100000
)

fg_time_df <- as.data.frame(fg_time)

 sig_time   <- fg_time_df$pathway[fg_time_df$padj < 0.05]

fg_time_plot <- fg_time_df[match(pathways_keep, fg_time_df$pathway), ]

fg_time_plot$order<-match(fg_time_plot$pathway,nes_mat$pathway)

fg_time_plot$pathway_clean <- gsub("HALLMARK_", "", fg_time_plot$pathway)


fg_time_plot$nesflip<-(-fg_time_plot$NES)

hallmarkbarmetonly <- ggplot(fg_time_plot,
                      aes(x = nesflip, y = reorder(pathway_clean, order), fill = nesflip)) +
  geom_col() +
      geom_text(
    data = subset(fg_time_plot, padj < 0.05),
    aes(label = "*",
        hjust = ifelse(nesflip > 0, -0.3, 1.3)),
    size = 6,
    fontface = "bold"
  ) +
  scale_x_reverse() +
  scale_y_discrete(expand = expansion(mult = c(0.1, 0.0))) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "gray95",
    high = "#b2182b",
    midpoint = 0,
        limits = c(-2.3, 4.3),
    name = "NES"
  ) +
  theme_minimal()  +
    annotate(
        "text",
        x = 0,
        y = 0,   # adjust depending on spacing
        label = "ŌåÉ Enriched Earlier     Enriched Later ŌåÆ",
        vjust = 1,  size = 3
    )+
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0)
  )+
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50")+
  labs(
    x = "",
    y = "",
    title = "Hallmark Pathway Enrichment Across Recurrence Time-\nJAVELIN Renal 101- Metastatic Site Sequencing Only"
  )+theme(legend.position = "none")


#primary only
#primary only
#primary only
#primary only
#primary only
#primary only
#primary only
finalexp<-subset(final,final$natid %in% rownames(tpmjav_z))
finalexp <- subset(finalexp, categories != "De Novo Metastatic")
finalexp<-subset(finalexp, site != "METASTASIS")
finalexp$timetotrial <- as.numeric(finalexp$timetotrial)


fit <- lm(wgii ~ timetotrial, data = finalexp)
finalexp$predwgii<-predict(fit,newdata = finalexp)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")

summary(fit)



p_label <- "p = 0.043"


wgiiprimonly <- ggplot(finalexp, aes(x = timetotrialcut, y = wgii,colour=site)) +
    geom_point(alpha = 0.7) +
  geom_line(aes(x=timetotrial,y=predwgii),linewidth = 1.2)+
    annotate("text",
             x = 40,
             y = 0.5,
             label = p_label,
             hjust = 0,
             size = 4) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"        ))+
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +
    labs(
        x = "Recurrence Time (months)",
        y = "WGII",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(0,0.7), xlim = c(0,120)) +
    theme_minimal()

wgiiprimonly



fit <- lm(cellcycle ~ timetotrial, data = finalexp)
finalexp$predcellcycle<-predict(fit,newdata = finalexp)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")

summary(fit)



p_label <- "p = 0.045"


cellcycleprimonly <- ggplot(finalexp, aes(x = timetotrialcut, y = cellcycle,colour=site)) +
    geom_point(alpha = 0.7) +
  geom_line(aes(x=timetotrial,y=predcellcycle),linewidth = 1.2)+
    annotate("text",
             x = 40,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"        ))+
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +
    labs(
        x = "Recurrence Time (months)",
        y = "Cell Cycle",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
    theme_minimal()

cellcycleprimonly

fit <- lm(angio ~ timetotrial, data = finalexp)
finalexp$predangio<-predict(fit,newdata = finalexp)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")

summary(fit)



p_label <- "p = 1e-5"


angioprimonly <- ggplot(finalexp, aes(x = timetotrialcut, y = angio,colour=site)) +
    geom_point(alpha = 0.7) +
  geom_line(aes(x=timetotrial,y=predangio),linewidth = 1.2)+
    annotate("text",
             x = 40,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"        ))+
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +
    labs(
        x = "Recurrence Time (months)",
        y = "Angio",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
    theme_minimal()

angioprimonly

fit <- lm(stroma ~ timetotrial, data = finalexp)
finalexp$predstroma<-predict(fit,newdata = finalexp)
cbind(
  coef(fit),
  confint(fit)
)

anova(fit, test= "LRT")

summary(fit)



p_label <- "p = 0.0076"


stromaprimonly <- ggplot(finalexp, aes(x = timetotrialcut, y = stroma,colour=site)) +
    geom_point(alpha = 0.7) +
  geom_line(aes(x=timetotrial,y=predstroma),linewidth = 1.2)+
    annotate("text",
             x = 40,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"        ))+
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +
    labs(
        x = "Recurrence Time (months)",
        y = "Stroma",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
    theme_minimal()

stromaprimonly



#hallmark
design <- model.matrix(~ timetotrial
                       
                       , data = finalexp)

ftpm<-tpmjav_num[match(finalexp$natid,rownames(tpmjav_z)),]


fit_time <- lmFit(t(ftpm), design)
fit_time <- eBayes(fit_time)


rank_time <- fit_time$t[, "timetotrial"]
names(rank_time) <- rownames(fit_time)

rank_time <- sort(rank_time, decreasing = TRUE)


fg_time <- fgsea(
  pathways = hallmark_list,
  stats = rank_time
  , nperm = 1000000
)

fg_time_df <- as.data.frame(fg_time)

 sig_time   <- fg_time_df$pathway[fg_time_df$padj < 0.05]

fg_time_plot <- fg_time_df[match(pathways_keep, fg_time_df$pathway), ]

fg_time_plot$order<-match(fg_time_plot$pathway,nes_mat$pathway)

fg_time_plot$pathway_clean <- gsub("HALLMARK_", "", fg_time_plot$pathway)


fg_time_plot$nesflip<-(-fg_time_plot$NES)

hallmarkbarprimonly <- ggplot(fg_time_plot,
                      aes(x = nesflip, y = reorder(pathway_clean, order), fill = nesflip)) +
  geom_col() +
      geom_text(
    data = subset(fg_time_plot, padj < 0.05),
    aes(label = "*",
        hjust = ifelse(nesflip > 0, -0.3, 1.3)),
    size = 6,
    fontface = "bold"
  ) +
  scale_x_reverse() +
  scale_y_discrete(expand = expansion(mult = c(0.1, 0.0))) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "gray95",
    high = "#b2182b",
    midpoint = 0,
        limits = c(-2.3, 4.4),
    name = "NES"
  ) +
  theme_minimal()  +
    annotate(
        "text",
        x = 0,
        y = 0,   # adjust depending on spacing
        label = "ŌåÉ Enriched Earlier     Enriched Later ŌåÆ",
        vjust = 1,  size = 3
    )+
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0)
  )+
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50")+
  labs(
    x = "",
    y = "",
    title = "Hallmark Pathway Enrichment Across Recurrence Time-\nJAVELIN Renal 101- Primary Tumor Sequencing Only"
  )+theme(legend.position = "none")

hallmarkbarprimonly

# Right column: 2x2 grid of the smaller plots
right_met <- plot_grid(
  angiometonly, cellcyclemetonly,
  stromametonly, wgiimetonly,
  ncol = 2, nrow = 2
)

# Combine: hallmark on left (spanning full height), 2x2 on right
combinemetsens <- plot_grid(
  hallmarkbarmetonly, right_met,
  ncol = 2,
  rel_widths = c(1, 1.2)
)

right_prim <- plot_grid(
  angioprimonly, cellcycleprimonly,
  stromaprimonly, wgiiprimonly,
  ncol = 2, nrow = 2
)

combineprimsens <- plot_grid(
  hallmarkbarprimonly, right_prim,
  ncol = 2,
  rel_widths = c(1, 1.2)
)

#sarcomatoid only
#sarcomatoid only
#sarcomatoid only
#sarcomatoid only
#sarcomatoid only
#sarcomatoid only

finalexp<-subset(final,final$natid %in% rownames(tpmjav_z))
finalexp <- subset(finalexp, categories != "De Novo Metastatic")
finalexp<-subset(finalexp, finalexp$sarc=="FALSE")
finalexp$timetotrial <- as.numeric(finalexp$timetotrial)
# 
# 
# fit <- lm(wgii ~ timetotrial, data = finalexp)
# finalexp$predwgii<-predict(fit,newdata = finalexp)
# cbind(
#   coef(fit),
#   confint(fit)
# )
# 
# anova(fit, test= "LRT")
# 
# summary(fit)
# 
# 
# 
# p_label <- "p = 0.062"
# 
# 
# wgiinosarc <- ggplot(finalexp, aes(x = timetotrialcut, y = wgii,colour=site)) +
#     geom_point(alpha = 0.7) +
#   geom_line(aes(x=timetotrial,y=predwgii),linewidth = 1.2)+
#     annotate("text",
#              x = 40,
#              y = 0.5,
#              label = p_label,
#              hjust = 0,
#              size = 4) +  
#   scale_color_manual(
#                  values = c(
#                      "KIDNEY" = "#4d9221",
#                      "METASTASIS" = "#c51b7d"        ))+
#     scale_x_continuous(
#         breaks = c(0,30,60,90,120),
#         labels = c("0","30","60","90",">120")
#     ) +
#     labs(
#         x = "Recurrence Time (months)",
#         y = "WGII",
#   colour = "Sequencing\nSite"
#     ) +
#     coord_cartesian(ylim = c(0,0.7), xlim = c(0,120)) +
#     theme_minimal()
# 
# wgiinosarc
# 
# fit <- lm(cellcycle ~ timetotrial, data = finalexp)
# finalexp$predcellcycle<-predict(fit,newdata = finalexp)
# cbind(
#   coef(fit),
#   confint(fit)
# )
# 
# anova(fit, test= "LRT")
# 
# summary(fit)
# 
# 
# 
# p_label <- "p = 0.003"
# 
# 
# cellcyclenosarc <- ggplot(finalexp, aes(x = timetotrialcut, y = cellcycle,colour=site)) +
#     geom_point(alpha = 0.7) +
#   geom_line(aes(x=timetotrial,y=predcellcycle),linewidth = 1.2)+
#     annotate("text",
#              x = 40,
#              y = 2.2,
#              label = p_label,
#              hjust = 0,
#              size = 4) +  
#   scale_color_manual(
#                  values = c(
#                      "KIDNEY" = "#4d9221",
#                      "METASTASIS" = "#c51b7d"        ))+
#     scale_x_continuous(
#         breaks = c(0,30,60,90,120),
#         labels = c("0","30","60","90",">120")
#     ) +
#     labs(
#         x = "Recurrence Time (months)",
#         y = "Cell Cycle",
#   colour = "Sequencing\nSite"
#     ) +
#     coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
#     theme_minimal()
# 
# cellcyclenosarc
# 
# fit <- lm(angio ~ timetotrial, data = finalexp)
# finalexp$predangio<-predict(fit,newdata = finalexp)
# cbind(
#   coef(fit),
#   confint(fit)
# )
# 
# anova(fit, test= "LRT")
# 
# summary(fit)
# 
# 
# 
# p_label <- "p = 0.008"
# 
# 
# angionosarc <- ggplot(finalexp, aes(x = timetotrialcut, y = angio,colour=site)) +
#     geom_point(alpha = 0.7) +
#   geom_line(aes(x=timetotrial,y=predangio),linewidth = 1.2)+
#     annotate("text",
#              x = 40,
#              y = 2.2,
#              label = p_label,
#              hjust = 0,
#              size = 4) +  
#   scale_color_manual(
#                  values = c(
#                      "KIDNEY" = "#4d9221",
#                      "METASTASIS" = "#c51b7d"        ))+
#     scale_x_continuous(
#         breaks = c(0,30,60,90,120),
#         labels = c("0","30","60","90",">120")
#     ) +
#     labs(
#         x = "Recurrence Time (months)",
#         y = "Angio",
#   colour = "Sequencing\nSite"
#     ) +
#     coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
#     theme_minimal()
# 
# angionosarc
# 
# fit <- lm(stroma ~ timetotrial, data = finalexp)
# finalexp$predstroma<-predict(fit,newdata = finalexp)
# cbind(
#   coef(fit),
#   confint(fit)
# )
# 
# anova(fit, test= "LRT")
# 
# summary(fit)
# 
# 
# 
# p_label <- "p = 0.0014"
# 
# stromanosarc <- ggplot(finalexp, aes(x = timetotrialcut, y = stroma,colour=site)) +
#     geom_point(alpha = 0.7) +
#   geom_line(aes(x=timetotrial,y=predstroma),linewidth = 1.2)+
#     annotate("text",
#              x = 40,
#              y = 2.2,
#              label = p_label,
#              hjust = 0,
#              size = 4) +  
#   scale_color_manual(
#                  values = c(
#                      "KIDNEY" = "#4d9221",
#                      "METASTASIS" = "#c51b7d"        ))+
#     scale_x_continuous(
#         breaks = c(0,30,60,90,120),
#         labels = c("0","30","60","90",">120")
#     ) +
#     labs(
#         x = "Recurrence Time (months)",
#         y = "Stroma",
#   colour = "Sequencing\nSite"
#     ) +
#     coord_cartesian(ylim = c(-3,3), xlim = c(0,120)) +
#     theme_minimal()
# 
# stromanosarc

#hallmark
design <- model.matrix(~ timetotrial+site
                       
                       , data = finalexp)

ftpm<-tpmjav_num[match(finalexp$natid,rownames(tpmjav_z)),]


fit_time <- lmFit(t(ftpm), design)
fit_time <- eBayes(fit_time)


rank_time <- fit_time$t[, "timetotrial"]
names(rank_time) <- rownames(fit_time)

rank_time <- sort(rank_time, decreasing = TRUE)


fg_time <- fgsea(
  pathways = hallmark_list,
  stats = rank_time, nperm = 1000000
)

fg_time_df <- as.data.frame(fg_time)

 sig_time   <- fg_time_df$pathway[fg_time_df$padj < 0.05]

fg_time_plot <- fg_time_df[match(pathways_keep, fg_time_df$pathway), ]

fg_time_plot$order<-match(fg_time_plot$pathway,nes_mat$pathway)

fg_time_plot$pathway_clean <- gsub("HALLMARK_", "", fg_time_plot$pathway)


fg_time_plot$nesflip<-(-fg_time_plot$NES)

hallmarkbarnosarc <- ggplot(fg_time_plot,
                      aes(x = nesflip, y = reorder(pathway_clean, order), fill = nesflip)) +
  geom_col() +
      geom_text(
    data = subset(fg_time_plot, padj < 0.05),
    aes(label = "*",
        hjust = ifelse(nesflip > 0, -0.3, 1.3)),
    size = 6,
    fontface = "bold"
  ) +
  scale_x_reverse() +
  scale_y_discrete(expand = expansion(mult = c(0.1, 0.0))) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "gray95",
    high = "#b2182b",
    midpoint = 0,
        limits = c(-2.3, 4.3),
    name = "NES"
  ) +
  theme_minimal()  +
    annotate(
        "text",
        x = 0,
        y = 0,   # adjust depending on spacing
        label = "ŌåÉ Enriched Earlier     Enriched Later ŌåÆ",
        vjust = 1,  size = 3
    )+
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0)
  )+
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50")+
  labs(
    x = "",
    y = "",
    title = "Hallmark Pathway Enrichment Across Recurrence Time-\nJAVELIN Renal 101- Excluding Sarcomatoid (n=319)"
  )+theme(legend.position = "none")


##high grade only (3/4)

##high grade only (3/4)
##high grade only (3/4)
##high grade only (3/4)
##high grade only (3/4)


finalexp<-subset(final,final$natid %in% rownames(tpmjav_z))
finalexp <- subset(finalexp, categories != "De Novo Metastatic")
finalexp<-subset(finalexp, finalexp$grade %in% c("GRADE 3","GRADE 4"))
finalexp$timetotrial <- as.numeric(finalexp$timetotrial)

#hallmark
design <- model.matrix(~ timetotrial+site
                       
                       , data = finalexp)

ftpm<-tpmjav_num[match(finalexp$natid,rownames(tpmjav_z)),]


fit_time <- lmFit(t(ftpm), design)
fit_time <- eBayes(fit_time)


rank_time <- fit_time$t[, "timetotrial"]
names(rank_time) <- rownames(fit_time)

rank_time <- sort(rank_time, decreasing = TRUE)


fg_time <- fgsea(
  pathways = hallmark_list,
  stats = rank_time, nperm = 1000000
)

fg_time_df <- as.data.frame(fg_time)

 sig_time   <- fg_time_df$pathway[fg_time_df$padj < 0.05]

fg_time_plot <- fg_time_df[match(pathways_keep, fg_time_df$pathway), ]

fg_time_plot$order<-match(fg_time_plot$pathway,nes_mat$pathway)

fg_time_plot$pathway_clean <- gsub("HALLMARK_", "", fg_time_plot$pathway)


fg_time_plot$nesflip<-(-fg_time_plot$NES)

hallmarkbarhighgrade <- ggplot(fg_time_plot,
                      aes(x = nesflip, y = reorder(pathway_clean, order), fill = nesflip)) +
  geom_col() +
      geom_text(
    data = subset(fg_time_plot, padj < 0.05),
    aes(label = "*",
        hjust = ifelse(nesflip > 0, -0.3, 1.3)),
    size = 6,
    fontface = "bold"
  ) +
  scale_x_reverse() +
  scale_y_discrete(expand = expansion(mult = c(0.1, 0.0))) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "gray95",
    high = "#b2182b",
    midpoint = 0,
        limits = c(-2.3, 4.3),
    name = "NES"
  ) +
  theme_minimal()  +
    annotate(
        "text",
        x = 0,
        y = 0,   # adjust depending on spacing
        label = "ŌåÉ Enriched Earlier     Enriched Later ŌåÆ",
        vjust = 1,  size = 3
    )+
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0)
  )+
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50")+
  labs(
    x = "",
    y = "",
    title = "Hallmark Pathway Enrichment Across Recurrence Time-\nJAVELIN Renal 101- High-Grade Only (ISUP/Fuhrman 3 or 4; n=220)"
  )+theme(legend.position = "none")

hallmarkbarhighgrade

#low grade only


finalexp<-subset(final,final$natid %in% rownames(tpmjav_z))
finalexp <- subset(finalexp, categories != "De Novo Metastatic")
finalexp<-subset(finalexp, finalexp$grade %in% c("GRADE 1","GRADE 2"))
finalexp$timetotrial <- as.numeric(finalexp$timetotrial)

#hallmark
design <- model.matrix(~ timetotrial+site
                       
                       , data = finalexp)

ftpm<-tpmjav_num[match(finalexp$natid,rownames(tpmjav_z)),]


fit_time <- lmFit(t(ftpm), design)
fit_time <- eBayes(fit_time)


rank_time <- fit_time$t[, "timetotrial"]
names(rank_time) <- rownames(fit_time)

rank_time <- sort(rank_time, decreasing = TRUE)


fg_time <- fgsea(
  pathways = hallmark_list,
  stats = rank_time, nperm = 1000000
)

fg_time_df <- as.data.frame(fg_time)

 sig_time   <- fg_time_df$pathway[fg_time_df$padj < 0.05]

fg_time_plot <- fg_time_df[match(pathways_keep, fg_time_df$pathway), ]

fg_time_plot$order<-match(fg_time_plot$pathway,nes_mat$pathway)

fg_time_plot$pathway_clean <- gsub("HALLMARK_", "", fg_time_plot$pathway)


fg_time_plot$nesflip<-(-fg_time_plot$NES)

hallmarkbarlowgrade <- ggplot(fg_time_plot,
                      aes(x = nesflip, y = reorder(pathway_clean, order), fill = nesflip)) +
  geom_col() +
      geom_text(
    data = subset(fg_time_plot, padj < 0.05),
    aes(label = "*",
        hjust = ifelse(nesflip > 0, -0.3, 1.3)),
    size = 6,
    fontface = "bold"
  ) +
  scale_x_reverse() +
  scale_y_discrete(expand = expansion(mult = c(0.1, 0.0))) +
  scale_fill_gradient2(
    low = "#2166ac",
    mid = "gray95",
    high = "#b2182b",
    midpoint = 0,
        limits = c(-2.3, 4.3),
    name = "NES"
  ) +
  theme_minimal()  +
    annotate(
        "text",
        x = 0,
        y = 0,   # adjust depending on spacing
        label = "ŌåÉ Enriched Earlier     Enriched Later ŌåÆ",
        vjust = 1,  size = 3
    )+
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0)
  )+
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50")+
  labs(
    x = "",
    y = "",
    title = "Hallmark Pathway Enrichment Across Recurrence Time-\nJAVELIN Renal 101- Low-Grade Only (ISUP/Fuhrman 1 or 2; n=117)"
  )+theme(legend.position = "none")

hallmarkbarlowgrade


combhisthallmark <- plot_grid(hallmarkbar,
  hallmarkbarnosarc, hallmarkbarhighgrade,
  hallmarkbarlowgrade,
  ncol = 2
)

```


```{r IMM151 molecular cluster analysis}


#random forest
#https://www.projectpro.io/recipes/perform-random-forest-r

immexp <- read_csv("/Users/smonda/Library/CloudStorage/OneDrive-MichiganMedicine/Research/M vs P/Final/IMM151_TPM.csv")

#already looks log2 tpm

immexp<-as.data.frame(immexp)
#remove na genes
immexp<-subset(immexp,!is.na(immexp$symbol))
#removes duplicated genes
immexp<-subset(immexp,!duplicated(immexp$symbol))

#row names to genes
rownames(immexp) <- immexp$symbol


#can only run this part once
immexp<- immexp[,!colnames(immexp) %in% c("symbol","gene_name","...1")]

#z-score transformed
immexpz<-data.frame(scale(data.frame(t(immexp))))

library(matrixStats)

immclin <- read_csv("/Users/smonda/Library/CloudStorage/OneDrive-MichiganMedicine/Research/M vs P/Final/IMM151_clinical.csv", 
                    skip = 3)

immclin<- subset(immclin,immclin$RNASEQ_SAMPLE_ID %in% colnames(immexp))
#10% most variable genes in IMM151
imm10genes<-names(sort(rowVars(as.matrix(immexp)),decreasing=TRUE)[1:(nrow(immexp)/10)])
imm10genes<-subset(imm10genes,imm10genes %in% colnames(immexpz))
sharedimm10genes<-intersect(colnames(tpmjav_z),imm10genes)

#then isolates to shared genes
#using javelin genes
immnmftarget<-as.factor(immclin$NMF_GROUP)

training<-immexpz[,sharedimm10genes]
#reorder to align with immclin
training<-training[immclin$RNASEQ_SAMPLE_ID,]

model<-randomForest(immnmftarget~.,data= training)

#OOB estimate of  error rate: 17.5%

#input data
rfinput<-tpmjav_z[,sharedimm10genes]
# rfinput<-tpm_z[,sharedimm10genes]

#as.data.frame(lapply(nexpz[,PMgenes], type.convert, as.is = TRUE))

nmf<-data.frame(predict(model,newdata=rfinput))

nmf$id<-rownames(rfinput)

# cancerhx$nmf<-nmf$predict.model..newdata...rfinput.[match(cancerhx$USUBJID,nmf$id)]

final$nmf<-nmf$predict.model..newdata...rfinput.[match(final$natid,nmf$id)]

finals<-subset(final,!is.na(final$nmf))

categories_plot<-finals %>%
    mutate(
        categories_plot = factor(
            categories_plot,
            levels = c("De Novo Metastatic","early","intermediate","late"),
            labels = c("De Novo Metastatic",
                       "Early Recurrent",
                       "Intermediate Recurrent",
                       "Late Recurrent")
        ),
        site = factor(site, levels = c("KIDNEY","METASTASIS"))
    ) %>%
    count(categories_plot, site, nmf) %>%
    group_by(categories_plot, site) %>%
    mutate(prop = n / sum(n)) %>%
    ggplot(
        aes(
            x = interaction(categories_plot, site),
            y = prop,
            fill = nmf
        )
    ) +
    geom_col(color = "black") +
    
    # Y axis as %
    scale_y_continuous(labels = label_percent()) +
    
    # Only show recurrence labels (remove KIDNEY/METASTASIS)
    scale_x_discrete(
        labels = function(x) {
            parts <- strsplit(x, "\\.")
            sapply(parts, function(p) p[1])
        }
    ) +
    
    labs(
        x = NULL,
        y = "Percentage",
        fill = "NMF subtype",
        title = "IMM151 Molecular Cluster Distribution by Recurrence Timing and Site in JAVELIN Renal 101 (n=599)"
    ) +
    
    # Group labels underneath (PRIMARY vs METASTATIC)
    annotate(
        "text",
        x = c(2.5, 6.5),
        y = -0.01,
        label = c("Primary Tumor Sequencing",
                  "Metastatic Site Sequencing"),
        vjust = 1,
        size = 3.5
    ) +
    
    # Allow drawing outside panel
    coord_cartesian(ylim = c(0, 1), clip = "off") +
    
    theme_minimal() +
    theme(
        axis.text.x = element_text(size = 10),
        plot.margin = ggplot2::margin(10, 10, 50, 10)
    )+ geom_vline(xintercept = 4.5, linewidth = 0.8, color = "black")+
  scale_fill_manual(
  values = scales::hue_pal()(7),  # keeps default-like colors
  breaks = c("1","2","3","4","5","6","7"),
  labels = c(
    "1. Angio/stromal",
    "2. Angio",
    "3. Compl./ox.",
    "4. T-effector/prolif.",
    "5. Prolif.",
    "6. Stromal/prolif.",
    "7. snoRNA"
  )
)

categories_plot

imm010plot<-imm010placeborec
imm010plot<-imm010plot[!is.na(imm010plot$NMF_SUBTYPE),]

imm010categories_plotplacebo<-imm010plot %>%
    mutate(
        categories_plot = factor(
            categories_plot,
            levels = c("early","intermediate"),
            labels = c(
                       "Early Recurrent",
                       "Intermediate Recurrent")),
                    NMF_SUBTYPE = factor(NMF_SUBTYPE, levels = c("1","2","3","4","5","6","7")
        )
    ) %>%
    count(categories_plot, NMF_SUBTYPE) %>%
    group_by(categories_plot) %>%
    mutate(prop = n / sum(n)) %>%
    ggplot(
        aes(
            x = interaction(categories_plot),
            y = prop,
            fill = NMF_SUBTYPE
        )
    ) +
    geom_col(color = "black") +
    
    # Y axis as %
    scale_y_continuous(labels = label_percent()) +
    
    # Only show recurrence labels (remove KIDNEY/METASTASIS)
    scale_x_discrete(
        labels = function(x) {
            parts <- strsplit(x, "\\.")
            sapply(parts, function(p) p[1])
        }
    ) +
    
    labs(
        x = NULL,
        y = "Percentage",
        fill = "NMF subtype",
        title = "IMM151 Molecular Cluster in IMmotion010 (placebo, n=87)"
    ) +
    

    
    # Allow drawing outside panel
    coord_cartesian(ylim = c(0, 1), clip = "off") +
    
    theme_minimal() +
    theme(
        axis.text.x = element_text(size = 10),
        plot.margin = ggplot2::margin(10, 10, 50, 10))+
  scale_fill_manual(
  values = scales::hue_pal()(7),  # keeps default-like colors
  breaks = c("1","2","3","4","5","6","7"),
  labels = c(
    "1. Angio/stromal",
    "2. Angio",
    "3. Compl./ox.",
    "4. T-effector/prolif.",
    "5. Prolif.",
    "6. Stromal/prolif.",
    "7. snoRNA"
  )
)



imm010categories_plotplacebo

imm010plot<-imm010atezorec
imm010plot<-imm010plot[!is.na(imm010plot$NMF_SUBTYPE),]

imm010categories_plotatezo<-imm010plot %>%
    mutate(
        categories_plot = factor(
            categories_plot,
            levels = c("early","intermediate"),
            labels = c(
                       "Early Recurrent",
                       "Intermediate Recurrent")),
                    NMF_SUBTYPE = factor(NMF_SUBTYPE, levels = c("1","2","3","4","5","6","7")
        )
    ) %>%
    count(categories_plot, NMF_SUBTYPE) %>%
    group_by(categories_plot) %>%
    mutate(prop = n / sum(n)) %>%
    ggplot(
        aes(
            x = interaction(categories_plot),
            y = prop,
            fill = NMF_SUBTYPE
        )
    ) +
    geom_col(color = "black") +
    
    # Y axis as %
    scale_y_continuous(labels = label_percent()) +
    
    # Only show recurrence labels (remove KIDNEY/METASTASIS)
    scale_x_discrete(
        labels = function(x) {
            parts <- strsplit(x, "\\.")
            sapply(parts, function(p) p[1])
        }
    ) +
    
    labs(
        x = NULL,
        y = "Percentage",
        fill = "NMF subtype",
        title = "IMM151 Molecular Cluster in IMmotion010 (atezo, n=76)"
    ) +
    

    
    # Allow drawing outside panel
    coord_cartesian(ylim = c(0, 1), clip = "off") +
    
    theme_minimal() +
    theme(
        axis.text.x = element_text(size = 10),
        plot.margin = ggplot2::margin(10, 10, 50, 10))+
  scale_fill_manual(
  values = scales::hue_pal()(7),  # keeps default-like colors
  breaks = c("1","2","3","4","5","6","7"),
  labels = c(
    "1. Angio/stromal",
    "2. Angio",
    "3. Compl./ox.",
    "4. T-effector/prolif.",
    "5. Prolif.",
    "6. Stromal/prolif.",
    "7. snoRNA"
  )
)

#create the same thing IMM010 data
imm010categories_plotatezo

```

## Establishing adjusted IMDC risk factors are can be included as continuous variable with linear effect in downstream analyses
```{r IMDC linear vs spline assessment}

#Frank Harrel RMS was used for this approach- plotting the HR CI 95% and assessing non-linearity with ANOVA

dd <- datadist(fpfs)
options(datadist = "dd")

fit_spline <- cph(Surv(as.numeric(AVALC), CNSR == "0") ~ 
                  categories + ARMCD + rcs(imdc,3),
                  data = fpfs,
                  x=TRUE, y=TRUE)



p_imdc <- Predict(fit_spline, imdc, fun = exp)

plot(p_imdc,
     xlab = "Modified IMDC (0ŌĆō4)",
     ylab = "Hazard Ratio",
     main = "Effect of Modified IMDC on PFS",
     lwd = 2)

anova(fit_spline)

```

## Prognostic Analysis- PFS is primary outcome, first - survival ~ recurrence_category + arm
second - survival ~ recurrence_category + arm + imdc

Then repeat for DSS and OS

You can also embed plots, for example:



```{r prognostic cox analysis}
#PFS primary
fpfs$ARMCD <- factor(fpfs$ARMCD)
fpfs$ARMCD <- relevel(fpfs$ARMCD, ref="Sunitinib")
fos$ARMCD <- factor(fos$ARMCD)
fos$ARMCD <- relevel(fos$ARMCD, ref = "Sunitinib")

#unadjusted-only arm
fpfs$categories <- factor(fpfs$categories)
fpfs$categories <- relevel(fpfs$categories, ref = "De Novo Metastatic")

foss<-fos[fpfs$ARMCD=="Sunitinib",]
fosa<-fos[fpfs$ARMCD!="Sunitinib",]
fpfss<-fpfs[fpfs$ARMCD=="Sunitinib",]
fpfsa<-fpfs[fpfs$ARMCD!="Sunitinib",]


coxpfs<-coxph(Surv(as.numeric(AVALC), CNSR == "0"
           ) ~              ARMCD+categories     ,
                    data = fpfs)
coxpfs
summary(coxpfs)

#adjusted- arm + imdc

coxadjpfs<-coxph(Surv(as.numeric(AVALC), CNSR == "0"
           ) ~              ARMCD+categories  +imdc   ,
                    data = fpfs)
coxadjpfs
summary(coxadjpfs)

#OS- secondary 1

#unadjusted-only arm
coxos<-coxph(Surv(as.numeric(AVALC), CNSR == "0"
           ) ~              ARMCD+categories     ,
                    data = fos)
coxos
summary(coxos)

#adjusted- arm + imdc

coxadjos<-coxph(Surv(as.numeric(AVALC), CNSR == "0"
           ) ~              ARMCD+categories  +imdc   ,
                    data = fos)
coxadjos
summary(coxadjos)


#DSS- secondary 2
#unadjusted-only arm
coxdss<-coxph(Surv(as.numeric(AVALC), CNSR == "0"& ADTHCAUS == "DISEASE PROGRESSION"
           ) ~              ARMCD+categories     ,
                    data = fos)
coxdss
summary(coxdss)

#adjusted- arm + imdc

coxadjdss<-coxph(Surv(as.numeric(AVALC), CNSR == "0"& ADTHCAUS == "DISEASE PROGRESSION"
           ) ~              ARMCD+categories  +imdc   ,
                    data = fos)
coxadjdss
summary(coxadjdss)


# Global test
#now global tests for the impact of category
# Reduced model (no categories)
cox_pfs_no_cat <- coxph(
  Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD,
  data = fpfs
)


anova(cox_pfs_no_cat, coxpfs, test = "LRT")

cox_pfs_no_cat_adj <- coxph(
  Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD + imdc,
  data = fpfs
)

anova(cox_pfs_no_cat_adj, coxadjpfs, test = "LRT")

cox_os_no_cat <- coxph(
  Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD,
  data = fos
)

anova(cox_os_no_cat, coxos, test = "LRT")

cox_os_no_cat_adj <- coxph(
  Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD + imdc,
  data = fos
)

anova(cox_os_no_cat_adj, coxadjos, test = "LRT")

cox_dss_no_cat <- coxph(
  Surv(as.numeric(AVALC),
       CNSR == "0" & ADTHCAUS == "DISEASE PROGRESSION") ~ ARMCD,
  data = fos
)

anova(cox_dss_no_cat, coxdss, test = "LRT")

cox_dss_no_cat_adj <- coxph(
  Surv(as.numeric(AVALC),
       CNSR == "0" & ADTHCAUS == "DISEASE PROGRESSION") ~ ARMCD + imdc,
  data = fos
)

anova(cox_dss_no_cat_adj, coxadjdss, test = "LRT")

```
#KM Curves for Prognostic Assessment
``` {r KM prognostic }


um_colors <- c(
  "#00274C",  # Michigan Blue
  "#FFCB05",  # Michigan Maize
  "#6CACE4",  # light blue accent
  "#9A3324"   # optional burgundy accent (nice contrast)
)


pfsprog <- 
  ggsurvplot(
  survfit(Surv(as.numeric(AVALC), CNSR == "0") ~ categories, data = fpfs),
  data = fpfs,
  risk.table = TRUE,
  xlab = "Time (months)",
  ylab = "Progression Free Survival",
  title = "Progression-Free Survival by Recurrence Status",
  legend.title = "",
  xlim = c(0, 60),
  break.time.by = 12,
  palette = um_colors,
  legend.labs = c(
    "De Novo Metastatic",
    "Early Recurrent",
    "Intermediate Recurrent",
    "Late Recurrent"
  ),
  surv.scale = "percent",
  ggtheme = theme_minimal(base_size = 14)
)

pfsprog

pfsprogs <-   ggsurvplot(
  survfit(Surv(as.numeric(AVALC), CNSR == "0") ~ categories, data = fpfss),
  data = fpfs,
  risk.table = TRUE,
  xlab = "Time (months)",
  ylab = "Progression Free Survival",
  title = "Progression-Free Survival by Recurrence Status\nSunitinib Arm (n=355)",
  legend.title = "",
  xlim = c(0, 60),
  break.time.by = 12,
  palette = um_colors,
  legend.labs = c(
    "De Novo Metastatic",
    "Early Recurrent",
    "Intermediate Recurrent",
    "Late Recurrent"
  ),
  surv.scale = "percent",
  ggtheme = theme_minimal(base_size = 14)
)

pfsprogs

pfsproga <-   ggsurvplot(
  survfit(Surv(as.numeric(AVALC), CNSR == "0") ~ categories, data = fpfsa),
  data = fpfs,
  risk.table = TRUE,
  xlab = "Time (months)",
  ylab = "Progression Free Survival",
  title = "Progression-Free Survival by Recurrence Status\nAvelumab+Axitinib Arm (n=361)",
  legend.title = "",
  xlim = c(0, 60),
  break.time.by = 12,
  palette = um_colors,
  legend.labs = c(
    "De Novo Metastatic",
    "Early Recurrent",
    "Intermediate Recurrent",
    "Late Recurrent"
  ),
  surv.scale = "percent",
  ggtheme = theme_minimal(base_size = 14)
)

pfsproga


osprog <- ggsurvplot(
  survfit(Surv(as.numeric(AVALC), CNSR == "0") ~ categories, data = fos),
  data = fos,
  risk.table = TRUE,
  xlab = "Time (months)",
  ylab = "Overall Survival",
  title = "Overall Survival by Recurrence Status\n(n=706)",
  legend.title = "",
  xlim = c(0, 60),
  break.time.by = 12,
  palette = um_colors,
  legend.labs = c(
    "De Novo Metastatic",
    "Early Recurrent",
    "Intermediate Recurrent",
    "Late Recurrent"
  ),
  surv.scale = "percent",
  ggtheme = theme_minimal(base_size = 14)
)
osprog

osproga <- ggsurvplot(
  survfit(Surv(as.numeric(AVALC), CNSR == "0") ~ categories, data = fosa),
  data = fos,
  risk.table = TRUE,
  xlab = "Time (months)",
  ylab = "Overall Survival",
  title = "Overall Survival by Recurrence Status Avelumab Arm",
  legend.title = "",
  xlim = c(0, 60),
  break.time.by = 12,
  palette = um_colors,
  legend.labs = c(
    "De Novo Metastatic",
    "Early Recurrent",
    "Intermediate Recurrent",
    "Late Recurrent"
  ),
  surv.scale = "percent",
  ggtheme = theme_minimal(base_size = 14)
)
osproga

osprogs <- ggsurvplot(
  survfit(Surv(as.numeric(AVALC), CNSR == "0") ~ categories, data = foss),
  data = fos,
  risk.table = TRUE,
  xlab = "Time (months)",
  ylab = "Overall Survival",
  title = "Overall Survival by Recurrence Status-Sunitinib Arm",
  legend.title = "",
  xlim = c(0, 60),
  break.time.by = 12,
  palette = um_colors,
  legend.labs = c(
    "De Novo Metastatic",
    "Early Recurrent",
    "Intermediate Recurrent",
    "Late Recurrent"
  ),
  surv.scale = "percent",
  ggtheme = theme_minimal(base_size = 14)
)
osprogs


dssprog <- ggsurvplot(
  survfit(Surv(as.numeric(AVALC), CNSR == "0"&fos$ADTHCAUS=="DISEASE PROGRESSION") ~ categories, data = fos),
  data = fos,
  risk.table = TRUE,
  xlab = "Time (months)",
  ylab = "Overall Survival",
  title = "Disease Specific Survival by Recurrence Status",
  legend.title = "",
  xlim = c(0, 60),
  break.time.by = 12,
  palette = um_colors,
  legend.labs = c(
    "De Novo Metastatic",
    "Early Recurrent",
    "Intermediate Recurrent",
    "Late Recurrent"
  ),
  surv.scale = "percent",
  ggtheme = theme_minimal(base_size = 14)
)

dssprog
pfs_full <- plot_grid(
  pfsprog$plot,
  pfsprog$table,
  ncol = 1,
  rel_heights = c(3, 1)
)

os_full <- plot_grid(
  osprog$plot,
  osprog$table,
  ncol = 1,
  rel_heights = c(3, 1)
)

dss_full <- plot_grid(
  dssprog$plot,
  dssprog$table,
  ncol = 1,
  rel_heights = c(3, 1)
)

pfss_full <- plot_grid(
  pfsprogs$plot,
  pfsprogs$table,
  ncol = 1,
  rel_heights = c(3, 1)
)

pfsa_full <- plot_grid(
  pfsproga$plot,
  pfsproga$table,
  ncol = 1,
  rel_heights = c(3, 1)
)

p_all <- plot_grid(
  pfss_full,
  pfsa_full,
  os_full,
    nrow = 1,
  align = "hv",
  axis = "tblr"
)

p_all

```


``` {r PFS HR subgroup analysis}

df <- data.frame(fpfs)
df$ARMCD <- factor(df$ARMCD, levels = c("Sunitinib","Avelumab + Axitinib"))

# Ensure expected types
df <- df %>%
  mutate(
    # event indicator: TRUE if event, FALSE if censored
    event = CNSR %in% c("0", 0),
    time  = as.numeric(AVALC),
    ARMCD = factor(ARMCD),
    categories = factor(categories,
                        levels = c("De Novo Metastatic","early","intermediate","late"))
  )

# >>>> OPTIONAL: set your desired reference arm here (edit "Control" as needed) <<<<
# df$ARMCD <- relevel(df$ARMCD, ref = "Control")

# Helper: fit Cox in a subset and extract HR (with CI) for each non-reference level of ARMCD
fit_subset <- function(dat, subgroup_label) {
  dat <- dat %>% filter(!is.na(time), !is.na(event), !is.na(ARMCD))
  if (nlevels(dat$ARMCD) < 2L || nrow(dat) == 0L) {
    return(tibble(subgroup = character(), arm = character(),
                  hr = numeric(), lo = numeric(), hi = numeric(),
                  p = numeric(), n = integer(), events = integer()))
  }
  fit <- coxph(Surv(time, event) ~ ARMCD, data = dat)
  s   <- summary(fit)
  
  # rows correspond to comparisons vs reference arm
  comp_rows <- rownames(s$coefficients)
  comp_arm  <- sub("^ARMCD", "", comp_rows)
  
  # confint for coxph returns log-scale; exponentiate to HR scale
  ci <- suppressMessages(confint(fit))
  tibble(
    subgroup = subgroup_label,
    arm      = comp_arm,
    hr       = as.numeric(exp(s$coefficients[, "coef"])),
    lo       = as.numeric(exp(ci[, 1])),
    hi       = as.numeric(exp(ci[, 2])),
    p        = as.numeric(s$coefficients[, "Pr(>|z|)"]),
    n        = nrow(dat),
    events   = sum(dat$event, na.rm = TRUE)
  )
}

# Define the sets: overall + each category
sets <- list(
  "All evaluated patients"        = df,
  "De Novo Metastatic"  = df %>% filter(categories == "De Novo Metastatic"),
  "Early"               = df %>% filter(categories == "early"),
  "Intermediate"            = df %>% filter(categories == "intermediate"),
  "Late"                = df %>% filter(categories == "late")
)

# Fit all
res <- bind_rows(Map(fit_subset, sets, names(sets)))

# Make pretty labels and plotting order
res <- res %>%
  mutate(
    line = paste0(subgroup, 
                  "   (n=", n, ", events=", events, ")"),
    line = factor(line, levels = rev(unique(line))) # top=overall
  )

# Axis limits (log scale) with a little padding
xmin <- max(min(res$lo, na.rm = TRUE), 0.05)
xmax <- max(res$hi, na.rm = TRUE) * 1.15

res <- res %>%
  mutate(
    hr_label = sprintf("%.2f (%.2fŌĆō%.2f)", hr, lo, hi)
  )

# Forest plot
forestplot<-ggplot(res, aes(y = line, x = hr)) +
    
    # HR text column on left
    geom_text(aes(x = 0.07, label = hr_label),
              hjust = 0,
              size = 3.5) +
    
    # CI + point
    geom_point(size = 2.5) +
    geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.18) +
    
    # Reference line
    geom_vline(xintercept = 1, linetype = "dashed") +
    annotate(
  "text",
  x = 0.88,
  y = 0.3,   # slightly below lowest row
  label = "Avelumab + Axitinib Better    <=>    Sunitinib Better",
  vjust =0,
  size = 3.5
)+
coord_cartesian(clip = "off") +
    # Log scale with extra room on left
    scale_x_log10(
        limits = c(0.07, 2),
        breaks = c(0.25, 0.5, 1, 1.5, 2),
        minor_breaks = NULL
    ) +
    
    labs(
        x = NULL,
        y = NULL,
        title = "Recurrence Subgroup Analysis for Progression-Free Survival",
    ) +
    
    theme_minimal(base_size = 12) +
    theme(
        panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0),
        axis.text.y = element_text(hjust = 0),
        plot.margin = ggplot2::margin(5.5, 30, 5.5, 30)# creates left padding
    )+
  geom_text(
  data = data.frame(),
  aes(x = 0.07, 
      y = length(levels(res$line)) + 0.4),
  label = "HR (95% CI)",
  hjust = 0,
  fontface = "bold",
  size = 3.8,
  inherit.aes = FALSE
)
  

fpfsil<-subset(fpfs,fpfs$categories %in% c("intermediate","late"))

fpfsde<-subset(fpfs,!fpfs$categories %in% c("intermediate","late"))

fpfsi<-subset(fpfs,fpfs$categories %in% c("intermediate"))

fpfse<-subset(fpfs,fpfs$categories %in% c("early"))

fpfsl<-subset(fpfs,fpfs$categories %in% c("late"))

fpfsd<-subset(fpfs,fpfs$categories %in% c("De Novo Metastatic"))



pfsil<-ggsurvplot(
  survfit(Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD, data = fpfsil),
  data = fpfsil,
  risk.table = TRUE,
    title="Intermediate + Late Recurrence (n=283)",
  xlab = "Time (months)",
  ylab = "Progression Free Survival",
    pval = TRUE,
  # title = "Progression-Free Survival by Recurrence Status",
  legend.title = "",
  xlim = c(0, 60),
  break.time.by = 12,
  palette = um_colors,
  legend.labs = c(
    "Sunitinib",
        "Avelumab + Axitinib"
  ),
  surv.scale = "percent",
  ggtheme = theme_minimal(base_size = 14)
)

pfsde<-ggsurvplot(
  survfit(Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD, data = fpfsde),
  data = fpfsde,
  risk.table = TRUE,
  title="De Novo + Early Recurrence (n=433)",
  xlab = "Time (months)",
  ylab = "Progression Free Survival",
  pval = TRUE,
  # title = "Progression-Free Survival by Recurrence Status",
  legend.title = "",
  xlim = c(0, 60),
  break.time.by = 12,
  palette = um_colors,
  legend.labs = c(
    "Sunitinib",
    "Avelumab + Axitinib"
      ),
  surv.scale = "percent",
  ggtheme = theme_minimal(base_size = 14)
)


pfse<-ggsurvplot(
  survfit(Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD, data = fpfse),
  data = fpfse,
  risk.table = TRUE,
  title="Early Recurrence (n=137)",
  xlab = "Time (months)",
  ylab = "Progression Free Survival",
    pval = TRUE,
  # title = "Progression-Free Survival by Recurrence Status",
  legend.title = "",
  xlim = c(0, 60),
  break.time.by = 12,
  palette = um_colors,
  legend.labs = c(
    "Sunitinib",
    "Avelumab + Axitinib"
      ),
  surv.scale = "percent",
  ggtheme = theme_minimal(base_size = 14)
)


pfsd<-ggsurvplot(
  survfit(Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD, data = fpfsd),
  data = fpfsd,
  risk.table = TRUE,
  title="De Novo Metastatic (n=296)",
  xlab = "Time (months)",
    pval = TRUE,
  ylab = "Progression Free Survival",
  # title = "Progression-Free Survival by Recurrence Status",
  legend.title = "",
  xlim = c(0, 60),
  break.time.by = 12,
  palette = um_colors,
  legend.labs = c(
    "Sunitinib",
    "Avelumab + Axitinib"
      ),
  surv.scale = "percent",
  ggtheme = theme_minimal(base_size = 14)
)

pfsi<-ggsurvplot(
  survfit(Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD, data = fpfsi),
  data = fpfsi,
  risk.table = TRUE,
  title="Intermediate Recurrence (n=139)",
  xlab = "Time (months)",
    pval = TRUE,
  ylab = "Progression Free Survival",
  # title = "Progression-Free Survival by Recurrence Status",
  legend.title = "",
  xlim = c(0, 60),
  break.time.by = 12,
  palette = um_colors,
  legend.labs = c(
    "Sunitinib",
    "Avelumab + Axitinib"
      ),
  surv.scale = "percent",
  ggtheme = theme_minimal(base_size = 14)
)


pfsl<-ggsurvplot(
  survfit(Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD, data = fpfsl),
  data = fpfsl,
  risk.table = TRUE,
    pval = TRUE,
  title="Late Recurrence (n=144)",
  xlab = "Time (months)",
  ylab = "Progression Free Survival",
  # title = "Progression-Free Survival by Recurrence Status",
  legend.title = "",
  xlim = c(0, 60),
  break.time.by = 12,
  palette = um_colors,
  legend.labs = c(
    "Sunitinib",
    "Avelumab + Axitinib"
      ),
  surv.scale = "percent",
  ggtheme = theme_minimal(base_size = 14)
)

pfssubsupp <- plot_grid(
  pfsd$plot,
    pfse$plot,
  pfsd$table,
   pfse$table,
  pfsi$plot,pfsl$plot,
    pfsi$table,
   pfsl$table,
  ncol = 2,
    rel_heights = c(2.5, 1.2,2.5,1.2)
)

pfssub <- plot_grid(
  pfsde$plot,
    pfsil$plot,
  pfsde$table,
   pfsil$table,
  ncol = 2,
    rel_heights = c(2.5, 1)
)
pfssubforest <- plot_grid(
  pfssub,
  forestplot,
  ncol = 1,
    rel_heights = c(2, 1.5),rel_widths = c(1,1.5)
)

pfssubsupforest <- plot_grid(
  pfssubsupp,
  forestplot,
  ncol = 1,
    rel_heights = c(5, 1.5),rel_widths = c(1,1.5)
)

pfssubsupforest
pfssubforest

```


#Interaction Assessment
``` {r interaction assessment }

# assess for interaction by using a global interaction test, an ANOVA between the two models

#----------------------------------------------------
#1) PRIMARY 
#1) primary analysis sensitivity Assessment of recurrence timing as a continous variable among only recurrent patients (1 df)

fpfsrecur<-subset(fpfs,!fpfs$categories %in% "De Novo Metastatic")

# Reduced model
cox_no_int <- coxph(
    Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD + timetotrial + imdc,
    data = fpfsrecur
)

# Full model
cox_int <- coxph(
    Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD * timetotrial + imdc,
    data = fpfsrecur
)

# Global interaction test
anova(cox_no_int, cox_int)
#----------------------------------------------------
#2) Secondary biologic grouping: Assessment of recurrence timing as a binary variable (denovo|early vs int|late)

fpfs$comb<-fpfs$categories %in% c("De Novo Metastatic","early")

# Reduced model
cox_no_int <- coxph(
    Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD + comb + imdc,
    data = fpfs
)

# Full model
cox_int <- coxph(
    Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD * comb + imdc,
    data = fpfs
)

# Global interaction test
anova(cox_no_int, cox_int)

#----------------------------------------------------
#3)  Interaction Analysis is Time to Recurrecne Continous (1 df)

# Reduced model
cox_no_int <- coxph(
    Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD + timetotrialinc + imdc,
    data = fpfs
)

# Full model
cox_int <- coxph(
    Surv(as.numeric(AVALC), CNSR == "0") ~ ARMCD * timetotrialinc + imdc,
    data = fpfs
)

# Global interaction test
anova(cox_no_int, cox_int) 



```

#ORR by recurrence category
``` {r ORR}

#ORR plot
# (Optional) set display order

finalorr<-subset(final,!final$orr %in% c("--redacted--","NOT EVALUABLE"))
cat_order  <- c( "De Novo Metastatic","early", "intermediate", "late")
orr_order  <- c( "COMPLETE RESPONSE", 
                "PARTIAL RESPONSE", "STABLE DISEASE", "PROGRESSIVE DISEASE")

df_prop <- finalorr %>%
  mutate(
    categories = factor(categories, levels = cat_order),
    orr        = factor(orr,        levels = orr_order)
  ) %>%
  count(categories, ARM, orr, name = "n") %>%
  group_by(categories, ARM) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()



# ORR (CR+PR) box (ymin at top of SD+PD, ymax at top of stack)
df_box <- df_prop %>%
  group_by(categories, ARM) %>%
  summarise(
    ymin = sum(prop[orr %in% c("STABLE DISEASE", "PROGRESSIVE DISEASE")]),
    ymax = ymin + sum(prop[orr %in% c("COMPLETE RESPONSE", "PARTIAL RESPONSE")]),
    .groups = "drop"
  ) %>%
  mutate(highlight = "ORR (CR + PR)")

# PD box (ymin at top of CR+PR+SD, ymax at end of PD)
df_box_pd <- df_prop %>%
  group_by(categories, ARM) %>%
  summarise(
    pd   = sum(prop[orr == "PROGRESSIVE DISEASE"]),
    ymin = 0,
    ymax = ymin + pd,
    .groups = "drop"
  ) %>%
  filter(pd > 0) %>%                        # skip zero-height rectangles
  mutate(highlight = "PD rate")
# --- Standard error for ORR and PD ---
df_counts <- df_prop %>%
  group_by(categories, ARM) %>%
  summarise(
    N     = sum(n),
    p_ORR = sum(n[orr %in% c("COMPLETE RESPONSE","PARTIAL RESPONSE")]) / N,
    p_PD  = sum(n[orr == "PROGRESSIVE DISEASE"]) / N,
    .groups = "drop"
  ) %>%
  mutate(
    se_ORR = sqrt(p_ORR * (1 - p_ORR) / N),
    se_PD  = sqrt(p_PD  * (1 - p_PD ) / N),
    
    # map ORR SE to the top-of-stack ORR box (same region as your ORR rectangle)
    ylow_ORR  = pmax(0, 1 - pmin(1, p_ORR + se_ORR)),
    yhigh_ORR = pmin(1, 1 - pmax(0, p_ORR - se_ORR)),
    
    # PD SE (two options):
    # 1) baseline-anchored (matches your current PD rectangle with ymin=0)
    ylow_PD_base  = pmax(0, p_PD - se_PD),
    yhigh_PD_base = pmin(1, p_PD + se_PD),
    
    # 2) top-of-stack (use if you change PD rectangle to sit on the PD segment)
    ylow_PD_top  = pmax(0, 1 - pmin(1, p_PD + se_PD)),
    yhigh_PD_top = pmin(1, 1 - pmax(0, p_PD - se_PD))
  )

# ---- Add the SE error bars to your existing plot ----

df_prop$categories <- factor(df_prop$categories,
                             levels = c("De Novo Metastatic","early","intermediate","late"),
                             labels = c("De Novo\nMetastatic",
                                        "Early\nRecurrent",
                                        "Intermediate\nRecurrent",
                                        "Late\nRecurrent"))
df_counts$categories <- factor(df_counts$categories,
                             levels = c("De Novo Metastatic","early","intermediate","late"),
                             labels = c("De Novo\nMetastatic",
                                        "Early\nRecurrent",
                                        "Intermediate\nRecurrent",
                                        "Late\nRecurrent"))

ORRPlot<-ggplot(
  df_prop,
  aes(x = categories, y = prop, fill = orr)
) +
  geom_col() +
  # ORR outline
  geom_rect(
    data = df_box,
    aes(
      xmin = as.numeric(categories) - 0.4,
      xmax = as.numeric(categories) + 0.4,
      ymin = ymin,
      ymax = ymax,
      color = highlight
    ),
    inherit.aes = FALSE,
    fill = NA,
    linewidth = 1.1,
    show.legend = TRUE
  ) +
  # PD outline (your current baseline version)
  geom_rect(
    data = df_box_pd,
    aes(
      xmin = as.numeric(categories) - 0.4,
      xmax = as.numeric(categories) + 0.4,
      ymin = ymin,
      ymax = ymax,
      color = highlight
    ),
    inherit.aes = FALSE,
    fill = NA,
    linewidth = 1.1,
    show.legend = TRUE
  ) +
  # ORR SE bars (top-of-stack region)
  geom_errorbar(
    data = df_counts %>% mutate(highlight = "ORR (CR + PR)"),
    aes(x = categories, ymin = ylow_ORR, ymax = yhigh_ORR, color = highlight),
    width = 0.15, linewidth = 0.9, inherit.aes = FALSE
  ) +
  # PD SE bars ŌĆö baseline-anchored to match your PD rectangle
  geom_errorbar(
    data = df_counts %>% mutate(highlight = "PD rate"),
    aes(x = categories, ymin = ylow_PD_base, ymax = yhigh_PD_base, color = highlight),
    width = 0.15, linewidth = 0.9, inherit.aes = FALSE
  ) +

  scale_y_continuous(labels = scales::label_percent(),breaks = seq(0, 1, 0.20)) +



  scale_color_manual(
    name = NULL,
    values = c("ORR (CR + PR)" = "black",
               "PD rate"        = "#d73027"),
    guide = guide_legend(
      override.aes = list(fill = NA, linewidth = 1.1),
      order = 2
    )
  ) +
  guides(fill = guide_legend(order = 1)) +
  labs(
    x = NULL, y = "Proportion of patients", fill = "Response",
    title = "Objective Response by Recurrence Timing and Treatment Arm"
  ) +
  facet_grid(. ~ ARM, switch = "x",
             labeller = labeller(
               ARM = as_labeller(c(
                 "Avelumab 10 mg/kg IV Q2W in a 6-week cycle + axitinib 5 mg PO BID" =
                   "Avelumab + Axitinib",
                 "Sunitinib 50 mg PO QD on Schedule 4/2" =
                   "Sunitinib"
               ))
             ))+
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    strip.placement = "outside",
    strip.background = element_blank(),
    panel.grid.major.x = element_blank(),   # remove vertical major gridlines
    panel.grid.minor.x = element_blank(),   # remove vertical minor gridlines
    panel.grid.minor.y = element_line(color = "gray20", linewidth = 0.3), # keep horizontal minor
    panel.grid.major.y = element_line(color = "gray20")   
  )+
  scale_fill_manual(
  values = c(
    "COMPLETE RESPONSE"    = "#1a9850",
    "PARTIAL RESPONSE"     = "#a6d96a",
    "STABLE DISEASE"       = "#ffffbf",
    "PROGRESSIVE DISEASE"  = "#f46d43"
  )
)+
  geom_vline(xintercept = 4.7, linewidth = 0.8, color = "black")

ORRPlot

```
``` {r site and late recurrence}


angiosite<-ggplot(final, aes(x = timetotrial, y = angio,colour=sitespecclean)) +
    geom_point(alpha = 0.7)  +
    scale_x_continuous(
        breaks = c(0,60,120,180,240,300),
        labels = c("0","60","120","180","240",300)
    ) +  
    # scale_color_manual(
    #     values = c(
    #         "KIDNEY" = "#4d9221",
    #         "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "Angiogenesis",
        colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,300)) +
    theme_minimal()+
    theme(legend.position="none")

cellcyclesite<-ggplot(final, aes(x = timetotrial, y = cellcycle,colour=sitespecclean)) +
    geom_point(alpha = 0.7)  +
    scale_x_continuous(
        breaks = c(0,60,120,180,240,300),
        labels = c("0","60","120","180","240",300)
    ) +  
    # scale_color_manual(
    #     values = c(
    #         "KIDNEY" = "#4d9221",
    #         "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "Cell Cycle",
        colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-3,3), xlim = c(0,300)) +
    theme_minimal()

site25years <- plot_grid(
  angiosite,
  cellcyclesite,
  ncol = 2,
    rel_heights = c(1, 1),rel_widths = c(1,1.3)
)




fit <- lm(g15 ~ timetotrial + site, data = final)
# fit <- lm(emt ~ timetotrial + site, data = final)

summary(fit)
cbind(
  coef(fit),
  confint(fit)
)

fit_reduced <- lm(g15 ~ site, data = final[!is.na(final$timetotrial),])
anova(fit_reduced, fit, test= "LRT")


final$predg15<-predict(fit,newdata = final)

# final$partial_angio <- resid(fit) + coef(fit)["timetotrial"] * final$timetotrial

p_label <- "p = 0.065"

summary(fit)

g15cont <- ggplot(final, aes(x = timetotrialcut, y = g15,colour=site)) +
    geom_point(alpha = 0.7) +geom_line(aes(x=timetotrial,y=predg15),linewidth = 1.2) +
    annotate("text",
             x = 40,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +  
    scale_color_manual(
        values = c(
            "KIDNEY" = "#4d9221",
            "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "15 Gene Score",
        colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-22,20), xlim = c(0,120)) +
    theme_minimal()



fit <- lm(emt ~ timetotrial + site, data = final)
# fit <- lm(emt ~ timetotrial + site, data = final)

summary(fit)
cbind(
  coef(fit),
  confint(fit)
)

fit_reduced <- lm(emt ~ site, data = final[!is.na(final$timetotrial),])
anova(fit_reduced, fit, test= "LRT")


final$predemt<-predict(fit,newdata = final)

# final$partial_angio <- resid(fit) + coef(fit)["timetotrial"] * final$timetotrial

p_label <- "p = 8e-7"

summary(fit)

emtcont <- ggplot(final, aes(x = timetotrialcut, y = emt,colour=site)) +
    geom_point(alpha = 0.7) +geom_line(aes(x=timetotrial,y=predemt),linewidth = 1.2) +
    annotate("text",
             x = 40,
             y = 2.2,
             label = p_label,
             hjust = 0,
             size = 4) +
    scale_x_continuous(
        breaks = c(0,30,60,90,120),
        labels = c("0","30","60","90",">120")
    ) +  
    scale_color_manual(
        values = c(
            "KIDNEY" = "#4d9221",
            "METASTASIS" = "#c51b7d"        ))+
    labs(
        x = "Recurrence Time (months)",
        y = "EMT Score",
        colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-30,30), xlim = c(0,120)) +
    theme_minimal()

emtcont


fit <- lm(emt ~ categories_model + site, data = final)
cbind(
  coef(fit),
  confint(fit)
)

final$predemtcat<-predict(fit,newdata = final)
summary(fit)
# Reduced model (no recurrence timing)
fit_reduced <- lm(emt ~ site, data = final)
anova(fit_reduced, fit, test= "LRT")




calipers <- data.frame(
  group1 = c("late","late","late"),
  group2 = c("De Novo Metastatic","early","intermediate"),
  p = c(0.0007,8.3e-05,0.00012)
) |>
  filter(p < 0.05)



calipers$y.position <- c(17.1,20.5,23.9)
calipers$label <- sprintf("p = %.2g", calipers$p)


emtcat<-ggplot(final, aes(x = categories_plot, y = emt)) +
      geom_beeswarm(
    aes(color = site),size = 1.5,alpha=0.7,method="center")+
  geom_line(aes(y=predemtcat,group=site,colour=site),linewidth = 1.2)+
    geom_boxplot(outlier.shape = NA, alpha = 0.3, width = 0.6) +

    stat_pvalue_manual(
        calipers,
        label = "label",
        tip.length = 0.01
    ) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"
                 ))+
    scale_x_discrete(
        labels = c(
            "De Novo Metastatic" = "DeNovo",
            "early" = "Early",
            "intermediate" = "Intermediate",
            "late" = "Late"
        )
    ) +
    labs(
        x = NULL,
        y = "EMT",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-40, 40)) +
    theme_minimal()



fit <- lm(g15 ~ categories_model + site, data = final)
cbind(
  coef(fit),
  confint(fit)
)

final$predg15cat<-predict(fit,newdata = final)
summary(fit)
# Reduced model (no recurrence timing)
fit_reduced <- lm(g15 ~ site, data = final)
anova(fit_reduced, fit, test= "LRT")




calipers <- data.frame(
  group1 = c("late","late","late"),
  group2 = c("De Novo Metastatic","early","intermediate"),
  p = c(0.34,0.039,0.0043)
) |>
  filter(p < 0.05)



calipers$y.position <- c(15.1,17.5)
calipers$label <- sprintf("p = %.2g", calipers$p)


g15cat<-ggplot(final, aes(x = categories_plot, y = g15)) +
      geom_beeswarm(
    aes(color = site),size = 1.5,alpha=0.7,method="center")+
  geom_line(aes(y=predg15cat,group=site,colour=site),linewidth = 1.2)+
    geom_boxplot(outlier.shape = NA, alpha = 0.3, width = 0.6) +

    stat_pvalue_manual(
        calipers,
        label = "label",
        tip.length = 0.01
    ) +  
  scale_color_manual(
                 values = c(
                     "KIDNEY" = "#4d9221",
                     "METASTASIS" = "#c51b7d"
                 ))+
    scale_x_discrete(
        labels = c(
            "De Novo Metastatic" = "DeNovo",
            "early" = "Early",
            "intermediate" = "Intermediate",
            "late" = "Late"
        )
    ) +
    labs(
        x = NULL,
        y = "15 Gene Score",
  colour = "Sequencing\nSite"
    ) +
    coord_cartesian(ylim = c(-22, 20)) +
    theme_minimal()

g15emtcomb <- plot_grid(
  g15cont,g15cat,
  emtcont,emtcat,
  ncol = 2,
    rel_heights = c(1, 1),rel_widths = c(1,1)
)



```
